---
title: "Competition for abiotic resources"
author: "Gaurav Kandlikar"
output: html_document
runtime: shiny
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(purrr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
```


```{r}
includeMarkdown("abiotic_rc_overview.Rmd")
```

```{r parameter descriptions, echo = F}
pars_vars <- c("$N_i$",
               "$R_j$",
               "$r_i$",
               "$k_{ij}$",
               "$c_{ij}$",
               "$m_i$",
               "$S_j$",
               "$a_j$")
descriptions <- c("Population size of consumer species $i$",
                 "Abundance of resource $j$",
                 "Maximum per-capita growth rate of consumer species $i$",
                 "Half-saturation constant of consumer $i$ for resource $j$ (i.e. the concentration of resource $j$ at which species $i$ grows at half of its maximum growth rate)",
                 "Consumption rate of resource $j$ by species $i$",
                 "Mortality rate of consumer species $i$",
                 "Supply point (or maximum resource level) of resource $j$",
                 "Rate at which resource $j$ is converted from its unusable to usable form")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```



```{r}
sidebarLayout(
  sidebarPanel(style = "overflow-y:scroll; max-height: 400px",
    # User defined resource growth rate -------
    h5("Resource supply rates"),
    sliderInput("S1", label = "S1",
                min = 0.01, max = 25, value = 12, step = .1),
    # User defined resource growth rate -------
    sliderInput("S2", label = "S2",
                min = 0.01, max = 25, value = 12, step = .1),

    # User defined resource growth rate -------
    h5("Consumers' max intrinsic growth rates"),
    sliderInput("r1", label = "r1",
                min = 0.01, max = 2, value = 1.6, step = .1),
    # User defined resource growth rate -------
    sliderInput("r2", label = "r2",
                min = 0.01, max = 2, value = 1, step = .1),
    
    # User defined resource uptake rates -------
    h5("Half saturation constants"),
    sliderInput("k11", label = "k11",
                min = 1, max = 30, value = 18, step = .1),
    sliderInput("k12", label = "k12",
                min = 1, max = 30, value = 4, step = .1),
    sliderInput("k21", label = "k21",
                min = 1, max = 30, value = 2, step = .1),
    sliderInput("k22", label = "k22",
                min = 1, max = 30, value = 14, step = .1),
    # User defined mortality rates -------
    h5("Mortality rates"),
    sliderInput("m1", label = "m1",
                min = 0.001, max = .5, value = .2, step = .01),
    sliderInput("m2", label = "m2",
                min = 0.001, max = .5, value = .2, step = .01),
    
    # User defined resource uptake rates -------
    h5("Resource uptake rates"),
    sliderInput("c11", label = "c11",
                min = 0.01, max = 1, value = .25, step = .01),
    sliderInput("c12", label = "c12",
                min = 0.01, max = 1, value = .08, step = .01),
    sliderInput("c21", label = "c21",
                min = 0.01, max = 1, value = .1, step = .01),
    sliderInput("c22", label = "c22",
                min = 0.01, max = 1, value = .2, step = .01),

    # User defined resource acquisition rates -------
    h5("Resource conversion rates"),
    sliderInput("a1", label = "a1",
                min = 0.001, max = 1, value = .5, step = .01),
    sliderInput("a2", label = "a2",
                min = 0.001, max = 1, value = .5, step = .01),

    # User defined N1, N2, and R -------
    h5("Initial population size of consumers"),
    column(width = 6, numericInput("N1", label = "N1",
                min = 1, max = 100, value = 10)),
    column(width = 6, numericInput("N2", label = "N2",
                min = 1, max = 100, value = 10)),
    h5("Initial size of resource pools"),
    column(width = 6,
           numericInput("R1", label = "R1",
                min = 1, max = 100, value = 20)),
    column(width = 6,
           numericInput("R2", label = "R2",
                min = 1, max = 100, value = 20)),
    # User defined time ---------
    numericInput("time", label = "Number of time steps to project",
                min = 1, max = 5000, value = 500)
    ),

  # Panel of plots -----
  mainPanel(
    renderPlot({plot_out()}, width = 600,
               height = 300),
    hr(),
    renderTable(Rstar_df()),
  )
)


tilman_params <- reactive({
  c(S1 = input$S1, S2 = input$S2, r1 = input$r1, r2 = input$r2,
    k11 = input$k11, k12 = input$k12, k21 = input$k21, k22 = input$k22,
    m1 = input$m1, m2 = input$m2,
    c11 = input$c11, c12 = input$c12, c21 = input$c21, c22 = input$c22,
    a1 = input$a1, a2 = input$a2)
})
tilman_init <- reactive({
  c(N1 = input$N1, N2 = input$N2, R1 = input$R1, R2 = input$R2)
})
time <- reactive({seq(from = 0, to = input$time, by=.1)})

Rstar_vec <- reactive({ecoevoapps::run_abiotic_comp_rstar(tilman_params())})
Rstar_df <- reactive({
  Rstar_vec <- Rstar_vec()
  data.frame(species = c("N1", "N2"),
             R1star = c(Rstar_vec["R11"], Rstar_vec["R21"]),
             R2star = c(Rstar_vec["R12"], Rstar_vec["R22"]))
  })

tilman_out <- reactive({
  to <- data.frame(deSolve::ode(func = ecoevoapps::tilman_comp_essential,
                          y = tilman_init(), parms=tilman_params(), times = time()))
  to %>% dplyr::mutate_if(is.numeric, round, digits = 3)
})

tilman_out_long <- reactive({
  pivot_longer(tilman_out(), c(R1,R2,N1,N2), "species")
})

tilman_out_R1R2 <- reactive({
  ggplot(tilman_out()) +
  geom_segment(data = Rstar_df(), 
               aes(x = R1star, xend = R1star, y = R2star, yend = Inf, color = species), size = 1.5) +
  geom_segment(data = Rstar_df(), 
               aes(x = R1star, xend = Inf, y = R2star, yend = R2star, color = species), size = 1.5) +
  geom_path(aes(x = R1, y = R2), size = 1) +
    geom_point(data = tail(tilman_out(), 1), aes(x = R1, y = R2), size = 4, stroke = 2, shape = 21) + 
    xlab("Resource 1") + ylab("Resource 2") + 
  scale_color_brewer(name = "Consumer\nspecies", palette = "Set1") +
  ecoevoapps::theme_apps() +
    theme(legend.position = "none")
})

tilman_out_N1N2 <- reactive({
  tilman_out_long <- tilman_out_long() %>% dplyr::filter(species %in% c("N1", "N2"))
  ggplot(tilman_out_long) +
  geom_line(aes(x = time, y = value, color = species), size = 2) +
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()
})

plot_out <- reactive({
  tilman_out_R1R2() + tilman_out_N1N2()
})


```


## References  

Tilman, D. 1980. [Resources: A graphical-mechanistic approach to competition and predation](http://max2.ese.u-psud.fr/epc/conservation/PDFs/HIPE/Tilman1980.pdf). The American Naturalist.

Tilman, D. 1982. Resource Competition and Community Structure. Princeton University Press, Princeton, New Jersey. 296 pages. 

Miller, T., et al. 2005. [A Critical Review of Twenty Years’ Use of the Resource‐Ratio Theory](https://www.jstor.org/stable/10.1086/428681?seq=1). The American Naturalist. 

------------
 
App maintained by Gaurav Kandlikar (gaurav.kandlikar@gmail.com)
