---
title: "Competitive species interactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Competitive-species-interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes various models of interspecific competition in the `ecoevoapps` package. 

The simplest model of interspecific competition in this package is the [Lotka Volterra] competition model:

$$\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1 + \alpha N_2}{K})$$

```{r setup}
library(ecoevoapps)

time <- seq(0, 100, 0.1)
init <- c(N1 = 70, N2 = 50)
params <- c(r1 = .1, r2 = .15, K1 = 500, K2 = 600, a = .5, b = .5)
lvcomp_simulation <- run_lvcomp_model(time = time, init = init, params = params)
```
This generates a dataframe with columns `time`, `N1`, and `N2`:

```{r}
head(lvcomp_simulation)
```

We can then use the `ecoevoapps` function `plot_lvcomp_time` to generate a trajectory of these populations over time:

```{r}
plot_lvcomp_time(lvcomp_simulation)
```
