---
title: "Differential equations in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential equations in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Behind the scenes, `ecoevoapps` uses the wonderful [`deSolve`](http://desolve.r-forge.r-project.org/) package to solve the differential equations in `R`. If you want to explore how this package works so that you can simulate the dynamics of your own favorite model or tweak existing models in `ecoevoapps`, read on!


To use `deSolve`, we first need to write a function that describes the differential equations, which we can then solve using the function `deSolve::ode()`. 

## Example 1 - Lotka-Volterra competition

If we want to run the Lotka-Volterra competition model in R, we first write a function using the syntax as follows: 

```{r}
# Hand-coding the Lotka-Volterra competition model
lotka_volterra_competition <- function(time, init, params) {
  with (as.list(c(time, init, params)), {
    # description of parameters
    # r1 = per capita growth rate of species 1
    # N1 = population size of species 1
    # K1 = carying capacity of species 1
    # a = relative per capita effect of species 2 on species 1
    # r2 = per capita growth rate of species 2
    # N2 = population size of species 2
    # K2 = carrying capacity of species 2
    # b = relative per capita effect of species 1 on species 2

    # Differential equations
    dN1 <- r1*N1*(1 - (N1 + a*N2)/K1)
    dN2 <- r2*N2*(1 - (N2 + b*N1)/K2)

    # Return dN1 and dN2
    return(list(c(dN1, dN2)))
  })
}
```

With this function defined, we can now define the parameter values, initial values, and time over which we want to run a simulation:

```{r}
# define vectors of initial population sizes, time, and parameters
# define initial population sizes for the two species
init <- c(N1 = 10, N2 = 30)
time <- seq(0,500, by = 0.1)
# define the parameter vector -- note that the names should match
# the parameter names we used in the lotka_volterra_competition function above
params <- c(r1 = .2, r2 = .1, K1 = 500, K2 = 600, a = .9, b = 1.1)
```

We are now ready to simulate the dynamics as follows:

```{r}
lv_out <- deSolve::ode(func = lotka_volterra_competition,
                       y = init, times = time, parms = params)
# Plot model output
plot(lv_out[,"N1"]~lv_out[,"time"], type = "l", ylim = c(0, 700))
lines(lv_out[,"N2"]~lv_out[,"time"], type = "l", col = 2)
```


## Example 2 - predator-prey dynamics

This example shows how the same approach can be used to simulate the Lotka-Volterra predator-prey model with logistic growth in the prey species:

```{r}
# Define the function
lv_predprey_logPrey <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate
    # e = conversion efficiency
    # d = predator death rate
    # K = carrying capacity of the prey

    dH = r*H*(1 - H/K) - (a*H*P)
    dP = e*(a*H*P) - d*P
    return(list(c(dH = dH, dP = dP)))

  })
}

# Define parameters and initial conditions
time <- seq(0, 250, 0.1)
init <- c(H = 10, P = 10)
params <- c(r = 1, K = 250, a = 0.1, e = 0.2, d = 0.25)

# Now, we are ready to run the mode:

predprey_out <- deSolve::ode(func = lv_predprey_logPrey,
                       y = init, times = time, parms = params)
# Plot model output
plot(predprey_out[,"H"]~predprey_out[,"time"], type = "l", ylim = c(0,16))
lines(predprey_out[,"P"]~predprey_out[,"time"], type = "l", col = 2)

```

## More resources

We also strongly recommend that you explore some deSolve tutorials to understand the workflow.  ([ex. 1](https://cran.r-project.org/web/packages/deSolve/vignettes/deSolve.pdf); [ex. 2](https://www.youtube.com/watch?v=dmfAgoLzhnk); there's also an [example](https://www.youtube.com/watch?v=zRlgvUuc9QY) specifically on the Lotka-Volterra predator-prey models!). The package also includes the function `deSolve::dede()`, which we use in `ecoevoapps` to simulate the delayed-differential model for logistic growth (see `run_logistic_model()`). 

