---
title: "Competition with Explicit Resource Dynamics"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
library(kableExtra)
```

This app simulates the dynamics of two consumer species ($P_1$ and $P_2$) that both require the same resource species ($H$) to grow. This allows us to implicitly model resource competition between the two consumer species, since an increase in $P_1$ results in a decreased growth rate of $P_2$ because of $P_1$'s suppression of resource $H$. 

Even this fairly straightforward scenario can be modeled in many ways. E.g. we can consider a resource species $H$ that, in the absence of either consumer, grows either exponentially (with no intrinsic limit to population growth), or logistically. We can also consider consumer species that have a [Type I or a Type II functional response ](https://en.wikipedia.org/wiki/Functional_response) for the resource. 
System of equations describing the most complex version of this model:  

\[
\begin{align}
\frac{dH}{dt} &= rH  \bigl(1-qH\bigr) - \frac{a_1HP_1}{1+a_1T_{h1}H} - \frac{a_2HP_2}{1+a_2T_{h2}H} \\
\\
\frac{dP_1}{dt} &= e_{1} \frac{a_1HP_1}{1+a_1T_{h1}H} - d_1P_1
\\
\\
\frac{dP2}{dt} &= e_2 \frac{a_2HP_2}{1+a_2T_{h2}H} - d_2P_2
\end{align}
\]


```{r parameter descriptions, echo=F}
pars_vars <- c("$H$", 
               "$P_i$", 
               "$r$", 
               "$q$", 
               "$a_i$", 
               "$T_{h_i}$", 
               "$e_i$", 
               "$d_i$")
descriptions <- c("Population size of the prey",
                 "Population size of predator $i$",
                 "Per capita growth rate of prey",
                 "Per capita prey self-limitation",
                 "Attack rate of predator $i$",
                 "Handling time of predator $i$",
                 "Conversion efficiency of predator $i$",
                 "Death rate of predator $i$")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

- Note that if $q =0$, prey has exponential growth. If $q>0$, then prey has logistic growth with carrying capacity $K = \frac{1}{q}$.  
- Also note that if $T_{h_{i}} = 0$, predator $i$ has a Type 1 functional response.  

```{r, echo=FALSE}
wellPanel(
  fluidRow(
    column(4, 
           ### Ask users for parameter values ----
           ## start with Prey params
           numericInput("H", label = "Initial population size of Prey", min = 1, value = 30),
           sliderInput("r", label = "Per capita growth rate of Prey (r)", min = .0001, max = 1.0, value = .2),
           sliderInput("q", label = "Prey self-limitation rate (q)", min = 0, max = 0.10, step = 0.0001, value = .0066),
           br(),
           hr(),
           numericInput("t", label = "Timesteps", min = 1, value = 1000),
           checkboxGroupInput("fxnrep", 
                              label = "Display plot of functional responses?",
                              choices = c("Yes" = "Yes"),
                              selected = NULL)
           
    ),
    
    column(4,
           ## now pred1 params + initial conditions
           numericInput("P1", label = "Initial population size of Predator 1", min = 1, value = 25),
           sliderInput("a1", label = "Predator 1 attack rate (a1)", min = .001, max = 1.0, value = .02),
           sliderInput("T_h1", label = "Predator 1 handling time (Th_1)", min = 0, max = 1.0, value = 0.1),
           sliderInput("e1", label = "Predator 1 conversion efficiency (e1)", min = .001, max = 1.0, value = 0.4),
           sliderInput("d1", label = "Per capita death rate of Predator 1 (d1)", min = .0001, max = 1.0, value = .1)
    ),
    
    column(4,
           ## now pred2 params + initial conditions
           numericInput("P2", label = "Initial population size of Predator 2", min = 1, value = 25),
           sliderInput("a2", label = "Predator 2 attack rate (a2)", min = .001, max = 1.0, value = .02),
           sliderInput("T_h2", label = "Predator 2 handling time (Th_2)", min = 0, max = 1.0, value = 0.1),
           sliderInput("e2", label = "Predator 2 conversion efficiency (e2)", min = .001, max = 1.0, value = 0.39),
           sliderInput("d2", label = "Per capita death rate of Predator 2 (d2)", min = .0001, max = 1.0, value = .1)
    )
  ))


fluidRow(
  column(12,
         renderPlot(plots_to_print())
  ))


# Set the initial population sizes

init <- reactive({c(H = input$H , P1 = input$P1, P2 = input$P2)})

# Set the parameter values

# description of parameters:
# r = per capita growth rate (prey)
# a = attack rate 
# T_h = handling time
# e = conversion efficiency
# d = predator death rate 

params <- reactive({c(r = input$r, q = input$q,
                      a1 = input$a1, T_h1 = input$T_h1, 
                      e1 = input$e1, d1 = input$d1, 
                      a2 = input$a2, T_h2 = input$T_h2, 
                      e2 = input$e2, d2 = input$d2)})

# Time over which to simulate model dynamics
time <- reactive({seq(0,input$t,by = .1)})



# Simulate the dynamics
out <- reactive({
  data.frame(ecoevoapps::run_biotic_competition_model(time = time(), 
                                                      init = init(), 
                                                      params = params()))
})
  
# Reshape the data for plotting
out_long <- reactive({
  pivot_longer(out(), c(H, P1, P2), "Population")
})


# Plots --------
## abundance through time
abund_plot <- reactive({(ggplot(out_long()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()) %>% originator() 
})

## functional responses of sp1 and sp2

fr_df <- reactive({
  fr_df <- data.frame(prey = 1:2000)
  fr_df <- fr_df %>% mutate(y1 = (input$a1 * prey)/(1 + input$a1 * input$T_h1 * prey),
                            y2 = (input$a2 * prey)/(1 + input$a2 * input$T_h2 * prey))
  fr_df
})

plot_fxnrep_sp1 <- reactive({
  if("Yes" %in% input$fxnrep){
    (ggplot(fr_df(), aes(x = prey, y = y1)) +
      geom_path(col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
      xlab("Prey Density") +
      ylab("Prey Consumed \nper Predator") +
      ecoevoapps::theme_apps()) %>% originator()
  }
})

plot_fxnrep_sp2 <- reactive({
  if("Yes" %in% input$fxnrep){
    (ggplot(fr_df(), aes(x = prey, y = y2)) +
      geom_path(col = brewer.pal(n = 3, name = "Set1")[3], size = 2) +
      xlab("Prey Density") +
      ylab("Prey Consumed \nper Predator") +
      ecoevoapps::theme_apps()) %>% originator()
  }
})

plot_fxnreps <- reactive({
  if("Yes" %in% input$fxnrep){
  wrap_plots(plot_fxnrep_sp1(), plot_fxnrep_sp2(), nrow = 2)
  }
})

# Make a list of plots and print out plots based on which ones were requested ----
plots_to_print <- reactive({
  if("Yes" %in% input$fxnrep){
    wrap_plots(abund_plot(), plot_fxnreps(), design = "AAB")
  } else {
    wrap_plots(abund_plot(), ncol = 1)
  }
  
})

   
```

## References 

- Armstrong and McGeehee, 1976. [Coexistence of species competing for shared resources](http://www-users.math.umn.edu/~mcgehee/publications/Armstrong1976b/Armstrong1976TheorPopBio9p317.pdf). Theoretical Population Biology.  
- Armstrong and McGeehee, 1980. [Competitive Exclusion](https://www.journals.uchicago.edu/doi/10.1086/283553). The American Naturalist.   
- Chesson, 1990. [MacArthur’s consumer–resource model](http://eebweb.arizona.edu/Faculty/chesson/Peter/Reprints/1990_MacArthur%27s%20consumer-resource%20model.pdf) Theoretical Population Biology.   
- Chesson and Kuang, 2008.  [The interaction between predation and competition.](https://www.nature.com/articles/nature07248) Nature. 

