---
title: "Paylaşılan biyotik kaynak için rekabet"
output: html_document
runtime: shiny
---

***[中文](https://ecoevoapps.shinyapps.io/biotic_resource_competition_ch) | [Español](https://ecoevoapps.shinyapps.io/biotic_resource_competition_es) |  [English](https://ecoevoapps.shinyapps.io/biotic_resource_competition) | [português](https://ecoevoapps.shinyapps.io/biotic_resource_competition_pt) | Turkish  ***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
library(kableExtra)

biotic_competition <- function(time, init, params) {
  with (as.list(c(time, init, params)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # q = 1/prey carrying capacity
    # a = attack rate
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate

    dH_dt = r*H*(1 - q*H) - (a1*H*P1)/(1 + a1*T_h1*H) - (a2*H*P2)/(1 + a2*T_h2*H)
    dP1_dt = e1*(a1*H*P1)/(1 + a1*T_h1*H) - d1*P1
    dP2_dt = e2*(a2*H*P2)/(1 + a2*T_h2*H) - d2*P2
    return(list(c(dH = dH_dt, dP1 = dP1_dt, dP2 = dP2_dt)))

  })
}

run_biotic_competition_model <- function(time, init, params) {
  deSolve::ode(func = biotic_competition,
                y = init, times = time, parms = params)
}


```

Bu aplikasyon büyümek için aynı biyotik kaynağa, ya da tüketilen türe (ava, $H$) ihtiyaç duyan iki tüketici türün ($P_1$ and $P_2$) dinamiğini gösterir. Dolayısıyla, iki tüketici türün rekabeti halinde, $P_1$'in artışı, $P_2$'nin büyüme oranını, $P_1$'in kaynakları tüketmesiyle azaltır.   

Bu basit varsayımlarla bile bahsedilen dinamik çok farklı şekilde modellenebilir (mesela, her iki tüketicinin yokluğunda tüketilen türün, H’nin, S veya J tipi popülasyon büyümesini değerlendirebiliriz). 

The models can also consider different relationships between the density of resources and rate at which they are attacked by consumers. This relationship is called the [functional response](https://en.wikipedia.org/wiki/Functional_response), and we consider two types of relationships. First, we consider a "Type 1 functional response", in which the rate at which consumers consume resources is a linear function of the resource density. The second relationship is the "Type II functional response", in which the amount of resources consumed is a saturating function of resource density. The idea of the Type II functional response was developed by C.S. Holling in the 1959 paper "[Some characteristics of simple types of predation and parasitism](https://hahana.soest.hawaii.edu/cmoreserver/summercourse/2010/documents/Holling_1959b.pdf)". 

Aşağıdaki denklem sistemi bu modelin karmaşık versiyonudur:

\[
\begin{align}
\frac{dH}{dt} &= rH  \bigl(1-qH\bigr) - \frac{a_1HP_1}{1+a_1T_{h1}H} - \frac{a_2HP_2}{1+a_2T_{h2}H} \\
\\
\frac{dP_1}{dt} &= e_{1} \frac{a_1HP_1}{1+a_1T_{h1}H} - d_1P_1
\\
\\
\frac{dP2}{dt} &= e_2 \frac{a_2HP_2}{1+a_2T_{h2}H} - d_2P_2
\end{align}
\]


```{r parameter descriptions, echo=F}
pars_vars <- c("$H$", 
               "$P_i$", 
               "$r$", 
               "$q$", 
               "$a_i$", 
               "$T_{h_i}$", 
               "$e_i$", 
               "$d_i$")
descriptions <- c("Population size of the prey",
                 "Population size of predator $i$",
                 "Per capita growth rate of prey",
                 "Per capita prey self-limitation",
                 "Attack rate of predator $i$",
                 "Handling time of predator $i$",
                 "Conversion efficiency of predator $i$",
                 "Death rate of predator $i$")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parametre/Değişkenler", "Tanımlar")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

- $q =0$ olduğu takdirde, gıda üstel büyüme gösterir. Eger $q>0$, gıda $K = \frac{1}{q}$ taşıma kapasitesiyle lojistik büyüme gösterir. 
- Eğer $T_{h_{i}} = 0$ ise, tüketici $i$ tip 1 fonksiyonel büyüme gösterir.

```{r, echo=FALSE}
wellPanel(
  fluidRow(
    column(4, 
           ### Ask users for parameter values ----
           ## start with Prey params
           numericInput("H", label = "Initial population size of Prey", min = 1, value = 30),
           sliderInput("r", label = "Per capita growth rate of Prey (r)", min = .0001, max = 1.0, value = .2),
           sliderInput("q", label = "Prey self-limitation rate (q)", min = 0, max = 0.10, step = 0.0001, value = .0066),
           br(),
           hr(),
           numericInput("t", label = "Timesteps", min = 1, value = 1000),
           checkboxGroupInput("fxnrep", 
                              label = "Display plot of functional responses?",
                              choices = c("Yes" = "Yes"),
                              selected = NULL)
           
    ),
    
    column(4,
           ## now pred1 params + initial conditions
           numericInput("P1", label = "Initial population size of Predator 1", min = 1, value = 25),
           sliderInput("a1", label = "Predator 1 attack rate (a1)", min = .001, max = 1.0, value = .02),
           sliderInput("T_h1", label = "Predator 1 handling time (Th_1)", min = 0, max = 1.0, value = 0.1),
           sliderInput("e1", label = "Predator 1 conversion efficiency (e1)", min = .001, max = 1.0, value = 0.4),
           sliderInput("d1", label = "Per capita death rate of Predator 1 (d1)", min = .0001, max = 1.0, value = .1)
    ),
    
    column(4,
           ## now pred2 params + initial conditions
           numericInput("P2", label = "Initial population size of Predator 2", min = 1, value = 25),
           sliderInput("a2", label = "Predator 2 attack rate (a2)", min = .001, max = 1.0, value = .02),
           sliderInput("T_h2", label = "Predator 2 handling time (Th_2)", min = 0, max = 1.0, value = 0.1),
           sliderInput("e2", label = "Predator 2 conversion efficiency (e2)", min = .001, max = 1.0, value = 0.39),
           sliderInput("d2", label = "Per capita death rate of Predator 2 (d2)", min = .0001, max = 1.0, value = .1)
    )
  ))


fluidRow(
  column(12,
         renderPlot(plots_to_print())
  ))


# Set the initial population sizes

init <- reactive({c(H = input$H , P1 = input$P1, P2 = input$P2)})

# Set the parameter values

# description of parameters:
# r = per capita growth rate (prey)
# a = attack rate 
# T_h = handling time
# e = conversion efficiency
# d = predator death rate 

params <- reactive({c(r = input$r, q = input$q,
                      a1 = input$a1, T_h1 = input$T_h1, 
                      e1 = input$e1, d1 = input$d1, 
                      a2 = input$a2, T_h2 = input$T_h2, 
                      e2 = input$e2, d2 = input$d2)})

# Time over which to simulate model dynamics
time <- reactive({seq(0,input$t, by = .1)})

# Simulate the dynamics
out <- reactive({ run_biotic_comp_model(time = time(), 
                                        init = init(), 
                                        params = params())
})
  
# Make a plot of abundance over time
abund_plot <- reactive({
  plot_biotic_comp_time(out())
})

## functional responses of the two consumers
plot_fxnreps <- reactive({
    wrap_plots(plot_functional_responses(params()), ncol = 1)
})

# Make a list of plots and print out plots based on which ones were requested ----
plots_to_print <- reactive({
  if("Yes" %in% input$fxnrep){
    wrap_plots(abund_plot(), plot_fxnreps(), design = "AAB")
  } else {
    wrap_plots(abund_plot(), ncol = 1)
  }
  
})


   
```

## Referanslar 

- Armstrong and McGeehee, 1976. [Coexistence of species competing for shared resources](http://www-users.math.umn.edu/~mcgehee/publications/Armstrong1976b/Armstrong1976TheorPopBio9p317.pdf). Theoretical Population Biology.  
- Armstrong and McGeehee, 1980. [Competitive Exclusion](https://www.journals.uchicago.edu/doi/10.1086/283553). The American Naturalist.   
- Chesson, 1990. [MacArthur’s consumer–resource model](http://eebweb.arizona.edu/Faculty/chesson/Peter/Reprints/1990_MacArthur%27s%20consumer-resource%20model.pdf) Theoretical Population Biology.   
- Chesson and Kuang, 2008.  [The interaction between predation and competition.](https://www.nature.com/articles/nature07248) Nature. 

-------------


```{r, results='asis', echo = F}
ecoevoapps::print_app_footer(language = "tr")
```
