---
title: "Competition with Explicit Resource Dynamics"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
library(kableExtra)
```


System of equations describing the most complex version of this model:  

```{r competition, include=FALSE}

# competition

competition <- function(time, init, params) {
  with (as.list(c(time, init, params)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # q = 1/prey carrying capacity
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H*(1 - q*H) - (a1*H*P1)/(1 + a1*T_h1*H) - (a2*H*P2)/(1 + a2*T_h2*H)
    dP1_dt = e1*(a1*H*P1)/(1 + a1*T_h1*H) - d1*P1
    dP2_dt = e2*(a2*H*P2)/(1 + a2*T_h2*H) - d2*P2
    return(list(c(dH = dH_dt, dP1 = dP1_dt, dP2 = dP2_dt)))
    
  })
}
```  


\[
\begin{eqnarray}
\frac{dH}{dt} = rH  \bigl(1-qH\bigr) - \frac{a_1HP_1}{1+a_1T_{h1}H} - \frac{a_2HP_2}{1+a_2T_{h2}H} \\
\\
\frac{dP_1}{dt} = e_{1} \frac{a_1HP_1}{1+a_1T_{h1}H} - d_1P_1
\\
\\
\frac{dP2}{dt} = e_2 \frac{a_2HP_2}{1+a_2T_{h2}H} - d_2P_2
\end{eqnarray}
\]


```{r parameter descriptions, echo=F}
pars_vars <- c("$H$", 
               "$P_i$", 
               "$r$", 
               "$q$", 
               "$a_i$", 
               "$T_{h_i}$", 
               "$e_i$", 
               "$d_i$")
descriptions <- c("Population size of the prey",
                 "Population size of predator $i$",
                 "Per capita growth rate of prey",
                 "Per capita prey self-limitation",
                 "Attack rate of predator $i$",
                 "Handling time of predator $i$",
                 "Conversion efficiency of predator $i$",
                 "Death rate of predator $i$")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

- Note that if $q =0$, prey has exponential growth. If $q>0$, then prey has logistic growth with carrying capacity $K = \frac{1}{q}$.  
- Also note that if $T_{h_{i}} = 0$, predator $i$ has a Type 1 functional response.  

```{r rosenzweig-macarthur, echo=FALSE}
wellPanel(
  fluidRow(
    column(4, 
           ### Ask users for parameter values ----
           ## start with Prey params
           numericInput("H", label = "Initial population size of Prey", min = 1, value = 30),
           sliderInput("r", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .2),
           sliderInput("q", label = "Inverse of prey carrying capacity", min = 0, max = 0.10, step = 0.0001, value = .0066),
           br(),
           br(),
           hr(),
           numericInput("t", label = "Timesteps", min = 1, value = 1000)
           
    ),
    
    column(4,
           ## now pred1 params + initial conditions
           numericInput("P1", label = "Initial population size of Predator 1", min = 1, value = 25),
           sliderInput("a1", label = "Predator 1 attack rate", min = .001, max = 1.0, value = .02),
           sliderInput("T_h1", label = "Predator 1 handling time", min = 0, max = 1.0, value = 0.1),
           sliderInput("e1", label = "Predator 1 conversion efficiency", min = .001, max = 1.0, value = 0.4),
           sliderInput("d1", label = "Per capita death rate of Predator 1", min = .0001, max = 1.0, value = .1)
    ),
    
    column(4,
           ## now pred2 params + initial conditions
           numericInput("P2", label = "Initial population size of Predator 2", min = 1, value = 25),
           sliderInput("a2", label = "Predator 2 attack rate", min = .001, max = 1.0, value = .02),
           sliderInput("T_h2", label = "Predator 2 handling time", min = 0, max = 1.0, value = 0.1),
           sliderInput("e2", label = "Predator 2 conversion efficiency", min = .001, max = 1.0, value = 0.39),
           sliderInput("d2", label = "Per capita death rate of Predator 2", min = .0001, max = 1.0, value = .1)
    )
  ))


fluidRow(
  column(12,
         renderPlot(plot())
         # renderPlot(np())
         # renderPlot(dabund_plot())
  ))


# Set the initial population sizes

init <- reactive({c(H = input$H , P1 = input$P1, P2 = input$P2)})

# Set the parameter values

# description of parameters:
# r = per capita growth rate (prey)
# a = attack rate 
# T_h = handling time
# e = conversion efficiency
# d = predator death rate 

params <- reactive({c(r = input$r, q = input$q, a1 = input$a1 , T_h1 = input$T_h1 , e1 = input$e1 , d1 = input$d1, 
                      a2 = input$a2 , T_h2 = input$T_h2 , e2 = input$e2 , d2 = input$d2)})

# Time over which to simulate model dynamics
time <- reactive({seq(0,input$t,by = .1)})



# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out <- reactive({data.frame(deSolve::ode(func = competition, y = init(), parms = params(), times = time()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long <- reactive({pivot_longer(out(), c(H, P1, P2), "Population")})


# Plots
plot <- reactive({(ggplot(out_long()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()) %>% originator() 
  })

# np <- reactive({ggplot(out()) + 
#                          geom_path(aes(x = H, y = P1), size = 2) + 
#                          xlab("Number of Prey") + 
#                          ylab("Number of Predators") + 
#     geom_segment(x = out_rm()$H[min(5, round(length(time_rm()))/50)], #did this min thing in case someone asks for very few timesteps
#                  y = out_rm()$P[min(5, round(length(time_rm()))/50)], 
#                  xend = out_rm()$H[min(5 + 1, round(length(time_rm())/50) + 1)], 
#                  yend = out_rm()$P[min(5 + 1, round(length(time_rm())/50) + 1)], 
#                  arrow = arrow(length = unit(0.1, "npc"))) +
#                          ecoevoapps::theme_apps() +
#                          stat_function(fun = function(x) (input$r_rm/input$a_rm)*(1 - x/input$K_rm)*(1 + input$a_rm * input$T_h_rm * x), col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
#                          geom_vline(xintercept = input$d_rm/(input$e_rm*input$a_rm - input$a_rm*input$d_rm*input$T_h_rm), col = brewer.pal(n = 3, name = "Set1")[1], size = 2)})


# pop_out_rm <- reactive({
#   pop_out <- out_rm()
#   pop_out$dH <- c(NA, diff(pop_out$H))
#   pop_out$dP <- c(NA, diff(pop_out$P))
#   # not using pgrs right now but could be useful to calculate
#   pop_out <- pop_out %>% mutate(pgrH = dH/H, pgrP = dP/P)
#   pop_out
# })
# 
# pop_out_long_rm <- reactive({ pop_out_rm() %>%
#   select(time, pgrH, pgrP) %>%
#   pivot_longer(c(pgrH, pgrP), "group") %>%
#   mutate(group = factor(group, levels = c("pgrH", "pgrP")))
# })
# 
# 
# # Plot dS, dI, dR over time
# dabund_plot_rm <- reactive({ ggplot(pop_out_long_rm()) +
#     geom_line(aes(x = time, y = value, color = group), size = 2) +
#     scale_color_brewer(palette = "Set1") +
#     ylab("Change in population size") +
#     ecoevoapps::theme_apps()
# })
   
```




