---
title: "Consumer Resource Interactions"
author: "Madeline Cowen and Rosa M. McGuire"
date: "6/8/2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
```


## Consumer- Resource Interactions {.tabset}

### Type I Functional Response

```{r consumer resource functions, include=FALSE}

# Type I Functional Response

lv_pred1 <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Logistic prey

logprey <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    # K = carrying capacity of the prey
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Type II Functional Response

lv_pred2 <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Rosenzweig-MacArthur Model

rm_predation <-function(time,init,params) {
  with (as.list(c(time,init,params)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # K = prey carrying capacity
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
    
  })
}
```



\[
\begin{eqnarray}
\frac{dH}{dt} = rH - aHP\\
\\
\frac{dP}{dt} = eaHP - dP
\end{eqnarray}
\]

Where:

* $H =$ Prey population size
* $P =$ Predator population size
* $r =$ per capita growth rate of Prey
* $a =$ Predator attack rate
* $e =$ conversion efficiency
* $d =$ Predator death rate  

```{r type 1 pred prey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_lv_pred1", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred1", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("e_lv_pred1", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 1),
    sliderInput("d_lv_pred1", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_lv_pred1", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_lv_pred1", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    
    
    numericInput("t_lv_pred1", label = "Timesteps", min = 1, value = 100)
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred1()),renderPlot(np_lvpred1()))

)

 # Set the initial population sizes

init_lv_pred1 <- reactive({c(H = input$H_lv_pred1 , P = input$P_lv_pred1)})

# Set the parameter values

    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 

pars_lv_pred1 <- reactive({c(r = input$r_lv_pred1, a = input$a_lv_pred1 , e = input$e_lv_pred1, d = input$d_lv_pred1)})

# Time over which to simulate model dynamics
time_lv_pred1 <- reactive({seq(0, input$t_lv_pred1, by = .1)})



# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out_lv_pred1 <- reactive({data.frame(deSolve::ode(func = lv_pred1, 
                                                  y = init_lv_pred1(), parms = pars_lv_pred1(), times = time_lv_pred1()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long_lv_pred1 <- reactive({pivot_longer(out_lv_pred1(), c(H,P), "Population")})


# Plots
plot_lvpred1 <- reactive({ggplot(out_long_lv_pred1()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()})


np_lvpred1 <- reactive({ggplot(out_lv_pred1()) + 
                         geom_path(aes(x = H, y = P), size = 2) + 
                         xlab("Number of Prey") + 
                         ylab("Number of Predators") + 
                         ecoevoapps::theme_apps() +
                         geom_hline(yintercept = input$r_lv_pred1/input$a_lv_pred1, col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
                         geom_vline(xintercept = input$d_lv_pred1/(input$e_lv_pred1*input$a_lv_pred1), col = brewer.pal(n = 3, name = "Set1")[1], size = 2) })


   
```

### Logistic Prey
 

\[
\begin{eqnarray}
\frac{dH}{dt} = rH  \biggl(1-\frac{H}{K}\biggr) - aHP \\
\\
\frac{dP}{dt} = eaHP - dP
\end{eqnarray}
\]

Where:

* $H =$ Prey population size
* $P =$ Predator population size
* $r =$ per capita growth rate of Prey
* $K =$ Prey carrying capacity
* $a =$ Predator attack rate
* $e =$ conversion efficiency
* $d =$ Predator death rate  

```{r logprey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, e, d, K
    
    sliderInput("r_logprey", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    numericInput("K_logprey", label = "Carrying capacity of Prey", min = 1, value = 100),
    sliderInput("a_logprey", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("e_logprey", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 1),
    sliderInput("d_logprey", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_logprey", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_logprey", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    
    
    numericInput("t_logprey", label = "Timesteps", min = 1, value = 100)
    
  ),  
  
 
  mainPanel(renderPlot(plot_logprey()), renderPlot(np_logprey()))
  
  #mainPanel(renderPlot(lvpred2_np()))
)

 # Set the initial population sizes

init_logprey <- reactive({c(H = input$H_logprey, P = input$P_logprey)})

# Set the parameter values

    # description of parameters:
    # r = per capita growth rate (prey)
    # K = carrying capacity of prey
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 

pars_logprey <- reactive({c(r = input$r_logprey, K = input$K_logprey, a = input$a_logprey, e = input$e_logprey, d = input$d_logprey)})

# Time over which to simulate model dynamics
time_logprey <- reactive({seq(0, input$t_logprey, by = .1)})



# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out_logprey <- reactive({data.frame(deSolve::ode(func = logprey, y = init_logprey(), parms = pars_logprey(), times = time_logprey()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long_logprey <- reactive({pivot_longer(out_logprey(), c(H,P), "Population")})


# Plots
plot_logprey <- reactive({ggplot(out_long_logprey()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()})

np_logprey <- reactive({ggplot(out_logprey()) + 
                         geom_path(aes(x = H, y = P), size = 2) + 
                         xlab("Number of Prey") + 
                         ylab("Number of Predators") + 
                         ecoevoapps::theme_apps() +
    geom_abline(intercept = input$r_logprey/input$a_logprey, slope = (-1 * input$r_logprey)/(input$a_logprey * input$K_logprey), col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
                         geom_vline(xintercept = input$d_logprey/(input$e_logprey*input$a_logprey), col = brewer.pal(n = 3, name = "Set1")[1], size = 2)})


   
```

### Type II Functional Response

\[
\begin{eqnarray}
\frac{dH}{dt} = rH - \frac{aH}{1+aT_hH} P\\
\\
\frac{dP}{dt} = eP \frac{aH}{1+aT_hH} - dP
\end{eqnarray}
\]

Where:

* $H =$ Prey population size
* $P =$ Predator population size
* $r =$ per capita growth rate of Prey
* $a =$ Predator attack rate
* $T_h =$ Predator handling time
* $e =$ conversion efficiency
* $d =$ Predator death rate  

```{r eruptions, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_lv_pred2", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred2", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("T_h_lv_pred2", label = "Predator handling time", min = 0.001, max = 1.0, value = 0.2),
    sliderInput("e_lv_pred2", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 1),
    sliderInput("d_lv_pred2", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_lv_pred2", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_lv_pred2", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    
    
    numericInput("t_lv_pred2", label = "Timesteps", min = 1, value = 100)
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred2()),renderPlot(np_lvpred2()))
  

)

 # Set the initial population sizes

init_lv_pred2 <- reactive({c(H = input$H_lv_pred2 , P = input$P_lv_pred2)})

# Set the parameter values

    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 

pars_lv_pred2 <- reactive({c(r = input$r_lv_pred2, a = input$a_lv_pred2 , T_h = input$T_h_lv_pred2 , e = input$e_lv_pred2 , d = input$d_lv_pred2 )})

# Time over which to simulate model dynamics
time_lv_pred2 <- reactive({seq(0,input$t_lv_pred2,by = .1)})


# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out_lv_pred2 <- reactive({data.frame(deSolve::ode(func = lv_pred2, y = init_lv_pred2(), parms = pars_lv_pred2(), times = time_lv_pred2()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long_lv_pred2 <- reactive({pivot_longer(out_lv_pred2(), c(H,P), "Population")})


# Plots
plot_lvpred2 <- reactive({ggplot(out_long_lv_pred2()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()})

np_lvpred2 <- reactive({ggplot(out_lv_pred2()) + 
                         geom_path(aes(x = H, y = P), size = 2) + 
                         xlab("Number of Prey") + 
                         ylab("Number of Predators") + 
                         ecoevoapps::theme_apps() +
                         geom_abline(intercept = input$r_lv_pred2/input$a_lv_pred2, slope = input$r_lv_pred2*input$T_h_lv_pred2, col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
                         geom_vline(xintercept = input$d_lv_pred2/(input$e_lv_pred2*input$a_lv_pred2 - input$a_lv_pred2*input$d_lv_pred2*input$T_h_lv_pred2), col = brewer.pal(n = 3, name = "Set1")[1], size = 2)})


   
```

### Rosenzweig- MacArthur Model 


\[
\begin{eqnarray}
\frac{dH}{dt} = rH  \biggl(1-\frac{H}{K}\biggr) - \frac{aHP}{1+aT_hH} \\
\\
\frac{dP}{dt} = e \frac{aHP}{1+aT_hH} - dP
\end{eqnarray}
\]

Where:

* $H =$ Prey population size
* $P =$ Predator population size
* $r =$ per capita growth rate of Prey
* $K =$ Prey carrying capacity
* $a =$ Predator attack rate
* $T_h =$ Predator handling time
* $e =$ conversion efficiency
* $d =$ Predator death rate  

```{r rosenzweig-macarthur, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_rm", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .2),
    numericInput("K_rm", label = "Prey carrying capacity", min = 1, value = 150),
    sliderInput("a_rm", label = "Predator attack rate", min = .001, max = 1.0, value = .02),
    sliderInput("T_h_rm", label = "Predator handling time", min = 0.001, max = 1.0, value = 1),
    sliderInput("e_rm", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 1),
    sliderInput("d_rm", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .4),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_rm", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("P_rm", label = "Initial population size of Predator", min = 1, value = 25),
    
    ### Ask users for time to simulate ----
    
    
    numericInput("t_rm", label = "Timesteps", min = 1, value = 100)
    
  ),  
  
 
  mainPanel(renderPlot(plot_rm()),renderPlot(np_rm()))
  
  #mainPanel(renderPlot(lvpred2_np()))
)

 # Set the initial population sizes

init_rm <- reactive({c(H = input$H_rm , P = input$P_rm)})

# Set the parameter values

    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 

params_rm <- reactive({c(r = input$r_rm, K = input$K_rm, a = input$a_rm , T_h = input$T_h_rm , e = input$e_rm , d = input$d_rm )})

# Time over which to simulate model dynamics
time_rm <- reactive({seq(0,input$t_rm,by = .1)})



# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out_rm <- reactive({data.frame(deSolve::ode(func = rm_predation, y = init_rm(), parms = params_rm(), times = time_rm()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long_rm <- reactive({pivot_longer(out_rm(), c(H,P), "Population")})


# Plots
plot_rm <- reactive({ggplot(out_long_rm()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") + 
  ecoevoapps::theme_apps()})

np_rm <- reactive({ggplot(out_rm()) + 
                         geom_path(aes(x = H, y = P), size = 2) + 
                         xlab("Number of Prey") + 
                         ylab("Number of Predators") + 
                         ecoevoapps::theme_apps() +
                         stat_function(fun = function(x) (input$r_rm/input$a_rm)*(1 - x/input$K_rm)*(1 + input$a_rm * input$T_h_rm * x), col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
                         geom_vline(xintercept = input$d_rm/(input$e_rm*input$a_rm - input$a_rm*input$d_rm*input$T_h_rm), col = brewer.pal(n = 3, name = "Set1")[1], size = 2)})


   
```


