---
title: "Consumer Resource Type I"
author: "Madeline Cowen"
output: html_document
runtime: shiny
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# library(tidyverse)
library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
```

## Consumer- Resource Interactions {.tabset}

### Type I Functional Response

```{r consumer resource functions, include=FALSE}

# Type I Functional Response

lv_pred1 <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Logistic prey

logprey <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    # K = carrying capacity of the prey
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

```
 

\[
\begin{eqnarray}
\frac{dH}{dt} = rH - aHP\\
\\
\frac{dP}{dt} = eaHP - dP
\end{eqnarray}
\]

Where:

* $H =$ Prey population size
* $P =$ Predator population size
* $r =$ per capita growth rate of Prey
* $a =$ Predator attack rate
* $e =$ conversion efficiency
* $d =$ Predator death rate  

```{r type 1 pred prey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_lv_pred1", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred1", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("e_lv_pred1", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 1),
    sliderInput("d_lv_pred1", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_lv_pred1", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_lv_pred1", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    
    
    numericInput("t_lv_pred1", label = "Timesteps", min = 1, value = 100)
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred1()),renderPlot(np_lvpred1()))

)

 # Set the initial population sizes

init_lv_pred1 <- reactive({c(H = input$H_lv_pred1 , P = input$P_lv_pred1)})

# Set the parameter values

    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 

pars_lv_pred1 <- reactive({c(r = input$r_lv_pred1, a = input$a_lv_pred1 , e = input$e_lv_pred1, d = input$d_lv_pred1)})

# Time over which to simulate model dynamics
time_lv_pred1 <- reactive({seq(0, input$t_lv_pred1, by = .1)})



# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out_lv_pred1 <- reactive({data.frame(deSolve::ode(func = lv_pred1, 
                                                  y = init_lv_pred1(), parms = pars_lv_pred1(), times = time_lv_pred1()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long_lv_pred1 <- reactive({pivot_longer(out_lv_pred1(), c(H,P), "Population")})


# Plots
plot_lvpred1 <- reactive({ggplot(out_long_lv_pred1()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()})

np_lvpred1 <- reactive({ggplot(out_lv_pred1()) + 
                         geom_path(aes(x = H, y = P), size = 2) + 
                         scale_color_brewer(palette = "Set1") + 
                         xlab("Number of Prey") + 
                         ylab("Number of Predators") + 
                         ecoevoapps::theme_apps()})


   
```

### Logistic Prey
 

\[
\begin{eqnarray}
\frac{dH}{dt} = rH  \biggl(1-\frac{H}{K}\biggr) - aHP \\
\\
\frac{dP}{dt} = eaHP - dP
\end{eqnarray}
\]

Where:

* $H =$ Prey population size
* $P =$ Predator population size
* $r =$ per capita growth rate of Prey
* $K =$ Prey carrying capacity
* $a =$ Predator attack rate
* $e =$ conversion efficiency
* $d =$ Predator death rate  

```{r logprey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, e, d, K
    
    sliderInput("r_logprey", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    numericInput("K_logprey", label = "Carrying capacity of Prey", min = 1, value = 100),
    sliderInput("a_logprey", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("e_logprey", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 1),
    sliderInput("d_logprey", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_logprey", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_logprey", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    
    
    numericInput("t_logprey", label = "Timesteps", min = 1, value = 100)
    
  ),  
  
 
  mainPanel(renderPlot(plot_logprey()), renderPlot(np_logprey()))
  
  #mainPanel(renderPlot(lvpred2_np()))
)

 # Set the initial population sizes

init_logprey <- reactive({c(H = input$H_logprey, P = input$P_logprey)})

# Set the parameter values

    # description of parameters:
    # r = per capita growth rate (prey)
    # K = carrying capacity of prey
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 

pars_logprey <- reactive({c(r = input$r_logprey, K = input$K_logprey, a = input$a_logprey, e = input$e_logprey, d = input$d_logprey)})

# Time over which to simulate model dynamics
time_logprey <- reactive({seq(0, input$t_logprey, by = .1)})



# Use the lv_competition function above to run 
# the lotka-volterra competition model using the 
# parameter estimates defined above

out_logprey <- reactive({data.frame(deSolve::ode(func = logprey, y = init_logprey(), parms = pars_logprey(), times = time_logprey()))})
  

# Reshape the data so that population sizes of both 
# species are in one column, and an extra column to define
# species name. This helps with the plotting...

out_long_logprey <- reactive({pivot_longer(out_logprey(), c(H,P), "Population")})


# Plots
plot_logprey <- reactive({ggplot(out_long_logprey()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
  ylab("Population size") +
  ecoevoapps::theme_apps()})

np_logprey <- reactive({ggplot(out_logprey()) + 
                         geom_path(aes(x = H, y = P), size = 2) + 
                         scale_color_brewer(palette = "Set1") + 
                         xlab("Number of Prey") + 
                         ylab("Number of Predators") + 
                         ecoevoapps::theme_apps()})


   
```
