---
title: "Consumer-Resource Dynamics"
author: "Madeline Cowen and Rosa M. McGuire"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
library(kableExtra)
ggplot2::theme_set(ecoevoapps::theme_apps())
```


This app presents several models that can be used to simulate the dynamics of consumers and biotic resources (e.g. predators and their prey, herbivores and plants, parasitoids and their hosts, etc. We refer to the resource as "prey" and the consumer as "predator". 

There are several variants of the general consumer-resource model that are presented here. The first tab presents the classic Lotka-Volterra predator prey model, and the other tabs all present variations on this classic model that add various aspects of biological reality. 


##  {.tabset}

### Lotka-Volterra predator-prey model

Equations describing prey that grows exponentially and predators that consume the prey following a Type I functional response:

\[
\begin{align}
\frac{dH}{dt} &= rH - aHP\\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params type 1, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a}\\
\\
H^* &= \frac{d}{ea} 
\end{align}
\]


```{r type 1 pred prey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_lv_pred1", label = "Per capita growth rate of Prey", 
                min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred1", label = "Predator attack rate", min = .001, 
                max = 1.0, value = .1),
    sliderInput("e_lv_pred1", label = "Predator conversion efficiency",
                min = .001, max = 1.0, value = 0.2),
    sliderInput("d_lv_pred1", label = "Per capita death rate of Predator",
                min = .0001, max = 1.0, value = .3),
    
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_lv_pred1", label = "Initial population size of Prey", 
                 min = 1, value = 10),
    numericInput("P_lv_pred1", label = "Initial population size of Predator", 
                 min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    numericInput("t_lv_pred1", label = "Timesteps", min = 10, value = 100),
    
    checkboxGroupInput("vectors_lv_pred1", label = "Display vector field?", 
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred1()),
            renderPlot(np_lvpred1())
            )

)

# Set the initial population sizes
init_lv_pred1 <- reactive({c(H = input$H_lv_pred1, P = input$P_lv_pred1)})

# Set the parameter values
pars_lv_pred1 <- reactive({
  c(r = input$r_lv_pred1, a = input$a_lv_pred1, 
    e = input$e_lv_pred1, d = input$d_lv_pred1)
})

# Time over which to simulate model dynamics
time_lv_pred1 <- reactive({seq(0, input$t_lv_pred1, by = .1)})

# simulate model dynamics
out_lv_pred1 <- reactive({
  data.frame(ecoevoapps::run_predprey_model(
    time = time_lv_pred1(),
    init = init_lv_pred1(),
    params = pars_lv_pred1()
  ))
})

# Plots ------
## make abundance thru time plot
plot_lvpred1 <- reactive({
  # pivot to a longer dataframe
  out_long_lv_pred1 <- pivot_longer(out_lv_pred1(), c(H,P), "Population")
  ggplot(out_long_lv_pred1) + 
    geom_line(aes(x = time, y = value, color = Population), size = 2) + 
    scale_color_brewer(palette = "Set1") +
    ylab("Population size") 
})

# plot trajectory of population
np_lvpred1 <- reactive({
  if("Yes" %in% input$vectors_lv_pred1) {
    ecoevoapps::plot_predprey_trajectory(out_lv_pred1(), pars_lv_pred1(), 
                                         vectors_field = TRUE)
  } else {
    ecoevoapps::plot_predprey_trajectory(out_lv_pred1(), pars_lv_pred1(), 
                             vectors_field = F)
  }
  
})

```

### Logistic Prey
 
This model presented here keeps the predator dynamics the same as in the L-V predator-prey model, but adds an additional layer of biological realism to the prey. Specifically, in the classic L-V model, the prey is modeled as having exponential growth, such that in the absence of predators, the prey grows without limits. In this tab, the prey experiences logistic growth, i.e. the prey grows logistically up to a carrying capacity $K$. 

\[
\begin{align}
\frac{dH}{dt} &= rH  \biggl(1-\frac{H}{K}\biggr) - aHP \\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params logistic growth, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$K$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of predator",
                 "Per capita growth rate of the prey",
                 "Carrying capacity of the prey",
                 "Attack rate of predator",
                 "Conversion efficiency of predator",
                 "Death rate of predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a}\bigg(1 - \frac{H}{K}\bigg)\\
\\
H^* &= \frac{d}{ea} \\
\end{align}
\]

```{r logprey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, e, d, K
    
    sliderInput("r_logprey", label = "Per capita growth rate of Prey",
                min = .0001, max = 1.0, value = .5),
    numericInput("K_logprey", label = "Carrying capacity of Prey", 
                 min = 1, value = 100),
    sliderInput("a_logprey", label = "Predator attack rate",
                min = .001, max = 1.0, value = .1),
    sliderInput("e_logprey", label = "Predator conversion efficiency", 
                min = .001, max = 1.0, value = 0.5),
    sliderInput("d_logprey", label = "Per capita death rate of Predator",
                min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    numericInput("H_logprey", label = "Initial population size of Prey",
                 min = 1, value = 10),
    numericInput("P_logprey", label = "Initial population size of Predator", 
                 min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    numericInput("t_logprey", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_logprey", label = "Display vector field?", 
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_logprey()), 
            renderPlot(np_logprey())
            )
)

# Set the initial population sizes
init_logprey <- reactive({
  c(H = input$H_logprey, P = input$P_logprey)
})

# Set the parameter values
pars_logprey <- reactive({
  c(r = input$r_logprey, K = input$K_logprey, 
    a = input$a_logprey, e = input$e_logprey, d = input$d_logprey)
  })

# Time over which to simulate model dynamics
time_logprey <- reactive({seq(0, input$t_logprey, by = .1)})

# Simulate the dynamics
out_logprey <- reactive({
    data.frame(ecoevoapps::run_predprey_model(
    time = time_logprey(),
    init = init_logprey(),
    params = pars_logprey()
  ))
})
  
# Plots ------

## make abundance thru time plot
plot_logprey <- reactive({
  out_long_logprey <- pivot_longer(out_logprey(), c(H,P), "Population")

  ggplot(out_long_logprey) + 
    geom_line(aes(x = time, y = value, color = Population), size = 2) + 
    scale_color_brewer(palette = "Set1") +
    ylab("Population size") 
})


np_logprey <- reactive({
  if("Yes" %in% input$vectors_logprey) {
    ecoevoapps::plot_predprey_trajectory(out_logprey(), pars_logprey(), 
                                         vectors_field = TRUE)
  } else {
    ecoevoapps::plot_predprey_trajectory(out_logprey(), pars_logprey(), 
                             vectors_field = FALSE)
  }
})

```

### Type II Functional Response

This model presented here keeps the predator dynamics the same as in the L-V predator-prey model, but adds an additional layer of biological realism to the *predator*. Specifically, in the classic L-V model, the predator has a *Type 1 functional response*, which means that predator individuals never get satiated, and don't spend any time handling the prey. In this model, the predator has a **Type II functional response**, which means that the predator's efficiency in eating the prey is limited by the time it has to invest in processing the prey. 

\[
\begin{align}
\frac{dH}{dt} &= rH - \frac{aH}{1+aT_hH} P\\
\\
\frac{dP}{dt} &= eP \frac{aH}{1+aT_hH} - dP
\end{align}
\]

```{r params type ii, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$T_{h}$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Handling time of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a} +rT_hH\\
\\
H^* &= \frac{d}{ea - adT_h} \\
\end{align}
\]


```{r eruptions, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    sliderInput("r_lv_pred2", label = "Per capita growth rate of Prey",
                min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred2", label = "Predator attack rate",
                min = .001, max = 1.0, value = .1),
    sliderInput("T_h_lv_pred2", label = "Predator handling time",
                min = 0.001, max = 1.0, value = 0.2),
    sliderInput("e_lv_pred2", label = "Predator conversion efficiency", 
                min = .001, max = 1.0, value = 0.7),
    sliderInput("d_lv_pred2", label = "Per capita death rate of Predator",
                min = .0001, max = 1.0, value = .3),
      
    ### Ask users for initial conditions -----
    numericInput("H_lv_pred2", label = "Initial population size of Prey",
                 min = 1, value = 10),
    numericInput("P_lv_pred2", label = "Initial population size of Predator",
                 min = 1, value = 10),
    
    ### Ask users for time to simulate ----    
    numericInput("t_lv_pred2", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_lv_pred2", label = "Display vector field?", 
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred2()),
            renderPlot(np_lvpred2())
  )
)

# Set the initial population sizes
init_lv_pred2 <- reactive({
  c(H = input$H_lv_pred2 , P = input$P_lv_pred2)
})

# Set the parameter values
pars_lv_pred2 <- reactive({
  c(r = input$r_lv_pred2, a = input$a_lv_pred2, 
    T_h = input$T_h_lv_pred2 , e = input$e_lv_pred2, 
    d = input$d_lv_pred2 )})

# Time over which to simulate model dynamics
time_lv_pred2 <- reactive({seq(0,input$t_lv_pred2,by = .1)})

# Simulate the dynamics
out_lv_pred2 <- reactive({
  data.frame(ecoevoapps::run_predprey_model(
    time = time_lv_pred2(),
    init = init_lv_pred2(),
    params = pars_lv_pred2()
  ))
})
  

# Plots ------

## make abundance thru time plot
plot_lvpred2 <- reactive({
  out_long_lv_pred2 <- pivot_longer(out_lv_pred2(), c(H,P), "Population")
  ggplot(out_long_lv_pred2) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
    ylab("Population size")
})

np_lvpred2 <- reactive({
  if("Yes" %in% input$vectors_lv_pred2) {
    ecoevoapps::plot_predprey_trajectory(out_lv_pred2(), pars_lv_pred2(), 
                                         vectors_field = TRUE)
  } else {
    ecoevoapps::plot_predprey_trajectory(out_lv_pred2(), pars_lv_pred2(), 
                             vectors_field = FALSE)
  }
})


```

### Rosenzweig-MacArthur Model 

The Rosenzweig-MacArthur model combines both logistic growth for the prey as well as a Type II functional response for the predator.

\[
\begin{align}
\frac{dH}{dt} &= rH  \biggl(1-\frac{H}{K}\biggr) - \frac{aHP}{1+aT_hH} \\
\\
\frac{dP}{dt} &= e \frac{aHP}{1+aT_hH} - dP
\end{align}
\]

```{r params rosen-mac, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$K$", 
               "$a$", 
               "$T_{h}$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Carrying capacity of the prey",
                 "Attack rate of the predator",
                 "Handling time of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a}\bigg(1 - \frac{H}{K}\bigg)(1 + aT_hH)\\
\\
H^* &= \frac{d}{ea - adT_h} \\
\end{align}
\]


```{r rosenzweig-macarthur, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_rm", label = "Per capita growth rate of Prey", 
                min = .0001, max = 1.0, value = .2),
    numericInput("K_rm", label = "Prey carrying capacity",
                 min = 1, value = 150),
    sliderInput("a_rm", label = "Predator attack rate",
                min = .001, max = 1.0, value = .02),
    sliderInput("T_h_rm", label = "Predator handling time",
                min = 0.001, max = 1.0, value = 0.3),
    sliderInput("e_rm", label = "Predator conversion efficiency", 
                min = .001, max = 1.0, value = 0.6),
    sliderInput("d_rm", label = "Per capita death rate of Predator",
                min = .0001, max = 1.0, value = .4),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    numericInput("H_rm", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("P_rm", label = "Initial population size of Predator", min = 1, value = 25),
    
    ### Ask users for time to simulate ----
    numericInput("t_rm", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_rm", label = "Display vector field?",
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),
  
  mainPanel(renderPlot(plot_rm()),
            renderPlot(np_rm())
            )
  
)

# Set the initial population sizes
init_rm <- reactive({c(H = input$H_rm , P = input$P_rm)})

# Set the parameter values
pars_rm <- reactive({
  c(r = input$r_rm, K = input$K_rm, 
    a = input$a_rm, T_h = input$T_h_rm, 
    e = input$e_rm, d = input$d_rm )})

# Time over which to simulate model dynamics
time_rm <- reactive({seq(0,input$t_rm,by = .1)})

out_rm <- reactive({
    data.frame(ecoevoapps::run_predprey_model(
    time = time_rm(),
    init = init_rm(),
    params = pars_rm()
  ))
  })
  
# Plots ------
## make abundance thru time plot
plot_rm <- reactive({
  out_long_rm <- pivot_longer(out_rm(), c(H,P), "Population")
  ggplot(out_long_rm) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
    ylab("Population size") 
})

np_rm <- reactive({
  if("Yes" %in% input$vectors_rm) {
    ecoevoapps::plot_predprey_trajectory(out_rm(), pars_rm(), 
                                         vectors_field = TRUE)
  } else {
    ecoevoapps::plot_predprey_trajectory(out_rm(), pars_rm(), 
                             vectors_field = FALSE)
  }
})

```

## Consumer Functional Response {.tabset}

The consumer's functional response captures the relationship between consumer's intake rate of the resource as a function of the resource density.

```{r functional response functions, echo=FALSE}
type1_fr <- function (H_init, H_final, a){
  to_return <- c(a*(H_init:H_final))
  return(to_return)
}

type2_fr <- function (H_init, H_final, a, Th){
  to_return <- c((a*(H_init:H_final)/(1+a*Th*(H_init:H_final))))
  return(to_return)
}

```

### Type I

Linear increase in intake rate with resource density, as shown by the equation:

$$f(H) = aH$$


```{r params type 1 fr, echo=F}
pars_vars <- c("$H$", 
               "$a$")
descriptions <- c("Population size of the prey",
                 "Attack rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  


```{r type 1 fr, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    # a
    sliderInput("a_fr_type1", label = "Predator attack rate", min = .001, max = 1.0, value = .02),
  
    ### Ask users for initial and final conditions -----
    #H_init, H_final
    
    numericInput("H_init_fr_type1", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("H_final_fr_type1", label = "Final population size of Prey", min = 1, value = 100), #Add warning: H_final>H_init 
    
  ),  
  
 mainPanel(renderPlot(plot_fr_type1())),

)


# Generate function output and data frame

out_fr_type1 <- reactive({type1_fr(H_init = input$H_init_fr_type1, 
                                   H_final = input$H_final_fr_type1,
                                   a = input$a_fr_type1)})

type1_fr_df <- reactive({data.frame(H = input$H_init_fr_type1:input$H_final_fr_type1,
                                    f_rate = out_fr_type1())})

#Plot

plot_fr_type1 <- reactive ({ggplot(type1_fr_df()) +
                            geom_line(aes(x = H, y=f_rate), size = 2) +
                            scale_color_brewer(palette = "Set1") +
                            xlab("Number of Prey") + 
                            ylab("Foraging Rate") +
                            ylim(0,50)})
   
```

### Type II

Decelerating intake rate, where the plateau represents consumer saturation:


$$g(H) = \frac {aH}{1+aT_hH}$$

```{r params type 2 fr, echo=F}
pars_vars <- c("$H$", 
               "$a$", 
               "$T_h$")
descriptions <- c("Population size of the prey",
                 "Attack rate of the predator",
                 "Handling time of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  


```{r type 2 fr, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    # a, Th
    
    sliderInput("a_fr_type2", label = "Predator attack rate", min = .001, max = 1.0, value = .02),
    sliderInput("T_h_fr_type2", label = "Predator handling time", min = 0.001, max = 1.0, value = 0.3),
  
    ### Ask users for initial and final conditions -----
    #H_init, H_final
    
    numericInput("H_init_fr_type2", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("H_final_fr_type2", label = "Final population size of Prey", min = 1, value = 100), #Add warning: H_final>H_init 
    
  ),  
  
 mainPanel(renderPlot(plot_fr_type2())),

)


# Generate function output and data frame

out_fr_type2 <- reactive({type2_fr(H_init = input$H_init_fr_type2, 
                                   H_final = input$H_final_fr_type2,
                                   a = input$a_fr_type2,
                                   Th = input$T_h_fr_type2)})

type2_fr_df <- reactive({data.frame(H = input$H_init_fr_type2:input$H_final_fr_type2,
                                    f_rate = out_fr_type2())})

#Plot

plot_fr_type2 <- reactive ({ggplot(type2_fr_df()) +
                            geom_line(aes(x = H, y=f_rate), size = 2) +
                            scale_color_brewer(palette = "Set1") +
                            xlab("Number of Prey") + 
                            ylab("Foraging Rate") +
                            ylim(0,4)})
   
```


## References 
- [Consumer Resource Dynamics](https://press.princeton.edu/books/paperback/9780691006574/consumer-resource-dynamics-mpb-36) by Murdoch, Briggs and Nisbet - digital copy may be available through a library.  
- [Graphical Representation and Stability Conditions of Predator-Prey Interactions](https://www.journals.uchicago.edu/doi/10.1086/282272), 1963, by M. L. Rosenzweig and R. H. MacArthur.   
- [Primer on functional response and numerical response](https://web.ma.utexas.edu/users/davis/375/popecol/lec10/funcresp.html)  
- [Some characteristics of simple types of predation and parasitism](https://hahana.soest.hawaii.edu/cmoreserver/summercourse/2010/documents/Holling_1959b.pdf), 1959, by C.S. Holling. 

---------

App maintained by Maddi Cowen (mcowen@g.ucla.edu) and Rosa McGuire (rmcguire1@ucla.edu)
