---
title: "Additional Consumer Resource Scenarios"
author: "Madeline Cowen"
date: "July 6, 2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
library(kableExtra)
```


## Consumer-Resource Variants and Scenarios {.tabset}



```{r consumer resource functions, include=FALSE}

## EQUATIONS FOR DESOLVE -------
# Type I Functional Response

lv_pred1 <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Logistic prey

logprey <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    # K = carrying capacity of the prey
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Type II Functional Response

lv_pred2 <- function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
  
  })
}

# Rosenzweig-MacArthur Model

rm_predation <-function(time,init,pars) {
  with (as.list(c(time,init,pars)), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # K = prey carrying capacity
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    return(list(c(dH = dH_dt, dP = dP_dt)))
    
  })
}

## EQ FOR SINGLE TIMESTEP CALCULATIONS -------

# exponential growth, type I
lv_pred1_eq <- function(H, P, pars) {
  
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
  
  })
}

# Logistic prey

logprey_eq <- function(H, P, pars) {
  
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    # K = carrying capacity of the prey
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
  
  })
}

# Type II Functional Response

lv_pred2_eq <- function(H, P, pars) {
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
  
  })
}

# Rosenzweig-MacArthur Model

rm_predation_eq <-function(H, P, pars) {
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # K = prey carrying capacity
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
    
  })
}

## make function to generate data for vector field/phase plane -------

vector_field_input <- function(sim_df, eq_func, pars_for_eq_func, vec_density = 20) {

  # INPUTS
  # sim_df is a df with the simulated values of H & P in separate columns
  # eq_func is the function with the system of equations that calculate dH and dP for one time step
  # pars_for_eq_func is the vector of named parameters to use in the eq_func
  # vec_density determines the number of arrows (vec_density^2)

  # OUTPUT:
  # a df with Hstart, Hend, Pstart, Pend (and dH, dP) for drawing vectors

  # BODY:
  # add error checks here

  # determine the min and max of the number of prey and predators
  lowH <- round(min(sim_df$H), 0)
  hiH <- round(max(sim_df$H), 0)
  lowP <- round(min(sim_df$P), 0)
  hiP <- round(max(sim_df$P), 0)

  # select a sequence of points between (and a little beyond) those values
  seqH <- seq(0.9*lowH, 1.4*hiH, length.out = vec_density)
  seqP <- seq(0.9*lowP, 1.4*hiP, length.out = vec_density)

  # find all the combinations of those H and P coordinates, make that a df w/Hstart and Pstart
  hpcoords <- expand.grid(Hstart = seqH, Pstart = seqP)

  # use those values to solve dP and dH and calculate pend and hend
  hpcoords <- bind_cols(hpcoords, map2_df(hpcoords$Hstart, hpcoords$Pstart, eq_func, pars_for_eq_func))
  hpcoords <- hpcoords %>% mutate(Hend = Hstart + dH, Pend = Pstart + dP)

  return(hpcoords)
}


# function to make the vector field with ggplot ------
vector_field <- function(sim_df, vector_field_input_data) {

  # INPUT
  # sim_df is a df with the simulated values of H & P in separate columns
  # vector_field_input_data has the output from vector_field_input, which is a list of start and end coordinates for each vector segment

  # OUTPUT
  # is a ggplot

  ggplot(sim_df) +

    # vector field
    geom_segment(data = vector_field_input_data,
                 aes(x = Hstart, y = Pstart, xend = Hend, yend = Pend),
                 arrow = arrow(length = unit(0.02, "npc")),
                 color = "light gray")

}


```

### Nicholson-Bailey Model (Discrete Time)

This is a classic host-parasitoid model in discrete time

**NOT THE CORRECT EQUATIONS, jUST LEAVING THEM HERE AS PLACEHOLDER**
\[
\begin{align}
\frac{dH}{dt} &= rH - aHP\\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params nic-bai, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

### Intraguild Predation

Equations describing prey that grows exponentially and predators that consume the prey following a Type I functional response:

**NOT THE CORRECT EQUATIONS, jUST LEAVING THEM HERE AS PLACEHOLDER**
\[
\begin{align}
\frac{dH}{dt} &= rH - aHP\\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params intraguild-pred, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

### Apparent Competition

**NOT THE CORRECT EQUATIONS, jUST LEAVING THEM HERE AS PLACEHOLDER**
\[
\begin{align}
\frac{dH}{dt} &= rH - aHP\\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params apparent-comp, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  
