---
title: "Consumer-Resource Dynamics"
author: "Madeline Cowen and Rosa M. McGuire"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(RColorBrewer)
library(kableExtra)
ggplot2::theme_set(ecoevoapps::theme_apps())
```


This app presents several models that can be used to simulate the dynamics of consumers and biotic resources (e.g. predators and their prey, herbivores and plants, parasitoids and their hosts, etc. We refer to the resource as "prey" and the consumer as "predator". 

There are several variants of the general consumer-resource model that are presented here. The first tab presents the classic Lotka-Volterra predator prey model, and the other tabs all present variations on this classic model that add various aspects of biological reality. 


##  {.tabset}


```{r consumer resource functions, include=FALSE}

## EQ FOR SINGLE TIMESTEP CALCULATIONS -------

# exponential growth, type I
lv_pred1_eq <- function(H, P, pars) {
  
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
  
  })
}

# Logistic prey

logprey_eq <- function(H, P, pars) {
  
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # e = conversion efficiency
    # d = predator death rate 
    # K = carrying capacity of the prey
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)
    dP_dt = e*(a*H*P) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
  
  })
}

# Type II Functional Response

lv_pred2_eq <- function(H, P, pars) {
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
  
  })
}

# Rosenzweig-MacArthur Model

rm_predation_eq <-function(H, P, pars) {
  with (as.list(pars), {
    # description of parameters:
    # r = per capita growth rate (prey)
    # K = prey carrying capacity
    # a = attack rate 
    # T_h = handling time
    # e = conversion efficiency
    # d = predator death rate 
    
    dH_dt = r*H*(1 - H/K) - (a*H*P)/(1 + a*T_h*H)
    dP_dt = e*(a*H*P)/(1 + a*T_h*H) - d*P
    
    return(data.frame(dH = dH_dt, dP = dP_dt))
    
  })
}

## make function to generate data for vector field/phase plane -------

vector_field_input <- function(sim_df, eq_func, pars_for_eq_func, vec_density = 20) {

  # INPUTS
  # sim_df is a df with the simulated values of H & P in separate columns
  # eq_func is the function with the system of equations that calculate dH and dP for one time step
  # pars_for_eq_func is the vector of named parameters to use in the eq_func
  # vec_density determines the number of arrows (vec_density^2)

  # OUTPUT:
  # a df with Hstart, Hend, Pstart, Pend (and dH, dP) for drawing vectors

  # BODY:
  # add error checks here

  # determine the min and max of the number of prey and predators
  lowH <- round(min(sim_df$H), 0)
  hiH <- round(max(sim_df$H), 0)
  lowP <- round(min(sim_df$P), 0)
  hiP <- round(max(sim_df$P), 0)

  # select a sequence of points between (and a little beyond) those values
  seqH <- seq(0.9*lowH, 1.4*hiH, length.out = vec_density)
  seqP <- seq(0.9*lowP, 1.4*hiP, length.out = vec_density)

  # find all the combinations of those H and P coordinates, make that a df w/Hstart and Pstart
  hpcoords <- expand.grid(Hstart = seqH, Pstart = seqP)

  # use those values to solve dP and dH and calculate pend and hend
  hpcoords <- bind_cols(hpcoords, map2_df(hpcoords$Hstart, hpcoords$Pstart, eq_func, pars_for_eq_func))
  hpcoords <- hpcoords %>% mutate(Hend = Hstart + dH, Pend = Pstart + dP)

  return(hpcoords)
}


# function to make the vector field with ggplot ------
vector_field <- function(sim_df, vector_field_input_data) {

  # INPUT
  # sim_df is a df with the simulated values of H & P in separate columns
  # vector_field_input_data has the output from vector_field_input, which is a list of start and end coordinates for each vector segment

  # OUTPUT
  # is a ggplot

  ggplot(sim_df) +

    # vector field
    geom_segment(data = vector_field_input_data,
                 aes(x = Hstart, y = Pstart, xend = Hend, yend = Pend),
                 arrow = arrow(length = unit(0.02, "npc")),
                 color = "light gray")

}


```

### Lotka-Volterra predator-prey model

Equations describing prey that grows exponentially and predators that consume the prey following a Type I functional response:

\[
\begin{align}
\frac{dH}{dt} &= rH - aHP\\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params type 1, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a}\\
\\
H^* &= \frac{d}{ea} 
\end{align}
\]


```{r type 1 pred prey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_lv_pred1", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred1", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("e_lv_pred1", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 0.2),
    sliderInput("d_lv_pred1", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    ### Ask users for initial conditions -----
    #N1, N2
    
    numericInput("H_lv_pred1", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_lv_pred1", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    numericInput("t_lv_pred1", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_lv_pred1", label = "Display vector field?", 
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred1()),
            renderPlot(np_lvpred1())
            # renderPlot(dabund_plot_lv_pred1())
            )

)

# Set the initial population sizes
init_lv_pred1 <- reactive({c(H = input$H_lv_pred1 , P = input$P_lv_pred1)})

# Set the parameter values
pars_lv_pred1 <- reactive({
  c(r = input$r_lv_pred1, a = input$a_lv_pred1 , e = input$e_lv_pred1, d = input$d_lv_pred1)
})

# Time over which to simulate model dynamics
time_lv_pred1 <- reactive({seq(0, input$t_lv_pred1, by = .1)})

# simulate model dynamics
out_lv_pred1 <- reactive({
  data.frame(ecoevoapps::run_predprey_model(
    time = time_lv_pred1(),
    init = init_lv_pred1(),
    params = pars_lv_pred1()
  ))
})
  
# Reshape the data for plotting
out_long_lv_pred1 <- reactive({
  pivot_longer(out_lv_pred1(), c(H,P), "Population")
})

# Plots ------
## make abundance thru time plot
plot_lvpred1 <- reactive({
  ggplot(out_long_lv_pred1()) + 
    geom_line(aes(x = time, y = value, color = Population), size = 2) + 
    scale_color_brewer(palette = "Set1") +
    ylab("Population size") 
})

## make isocline plot and add vector field depending on user input
vec_input_lv_pred1 <- reactive({
  vector_field_input(sim_df = out_lv_pred1(), eq_func = lv_pred1_eq,
                     pars_for_eq_func = pars_lv_pred1())
})

np_lvpred1 <- reactive({
  if("Yes" %in% input$vectors_lv_pred1) {
    vector_field(out_lv_pred1(), vec_input_lv_pred1()) +
      
      # the isoclines
      geom_hline(yintercept = input$r_lv_pred1/input$a_lv_pred1, 
                 col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
      geom_vline(xintercept = input$d_lv_pred1/(input$e_lv_pred1*input$a_lv_pred1),
                 col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
      
      # the trace of the simulation and arrow for direction of the trace
      geom_path(aes(x = H, y = P), size = 2) +
      geom_segment(x = out_lv_pred1()$H[min(5, round(length(time_lv_pred1()))/50)], #did this min thing in case someone asks for very few timesteps
                   y = out_lv_pred1()$P[min(5, round(length(time_lv_pred1()))/50)],
                   xend = out_lv_pred1()$H[min(5 + 1, round(length(time_lv_pred1())/50) + 1)],
                   yend = out_lv_pred1()$P[min(5 + 1, round(length(time_lv_pred1())/50) + 1)],
                   arrow = arrow(length = unit(0.1, "npc")),
                   cex = 2) +
      
      # plot appearance
      xlab("Number of Prey") +
      ylab("Number of Predators") +
      coord_cartesian(xlim = c(min(out_lv_pred1()$H), max(out_lv_pred1()$H) + 1), 
                      ylim = c(min(out_lv_pred1()$P), max(out_lv_pred1()$P) + 1))  #need this line to show all vectors that go beyond plot limits
    
  } else {
    
    ggplot(out_lv_pred1()) +
      # the isoclines
      geom_hline(yintercept = input$r_lv_pred1/input$a_lv_pred1, 
                 col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
      geom_vline(xintercept = input$d_lv_pred1/(input$e_lv_pred1*input$a_lv_pred1), 
                 col = brewer.pal(n = 3, name = "Set1")[2], size = 2) +
      
      # the trace of the simulation and arrow for direction of the trace
      geom_path(aes(x = H, y = P), size = 2) +
      geom_segment(x = out_lv_pred1()$H[min(5, round(length(time_lv_pred1()))/50)], #did this min thing in case someone asks for very few timesteps
                   y = out_lv_pred1()$P[min(5, round(length(time_lv_pred1()))/50)],
                   xend = out_lv_pred1()$H[min(5 + 1, round(length(time_lv_pred1())/50) + 1)],
                   yend = out_lv_pred1()$P[min(5 + 1, round(length(time_lv_pred1())/50) + 1)],
                   arrow = arrow(length = unit(0.1, "npc")),
                   cex = 2) +
      
      # plot appearance
      xlab("Number of Prey") +
      ylab("Number of Predators") +
      coord_cartesian(xlim = c(min(out_lv_pred1()$H), max(out_lv_pred1()$H) + 1),
                      ylim = c(min(out_lv_pred1()$P), max(out_lv_pred1()$P) + 1)) #need this line to show all vectors that go beyond plot limits

  }
  
})

```

### Logistic Prey
 
This model presented here keeps the predator dynamics the same as in the L-V predator-prey model, but adds an additional layer of biological realism to the prey. Specifically, in the classic L-V model, the prey is modeled as having exponential growth, such that in the absence of predators, the prey grows without limits. In this tab, the prey experiences logistic growth, i.e. the prey grows logistically up to a carrying capacity $K$. 

\[
\begin{align}
\frac{dH}{dt} &= rH  \biggl(1-\frac{H}{K}\biggr) - aHP \\
\\
\frac{dP}{dt} &= eaHP - dP
\end{align}
\]

```{r params logistic growth, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$K$", 
               "$a$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of predator",
                 "Per capita growth rate of the prey",
                 "Carrying capacity of the prey",
                 "Attack rate of predator",
                 "Conversion efficiency of predator",
                 "Death rate of predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a}\bigg(1 - \frac{H}{K}\bigg)\\
\\
H^* &= \frac{d}{ea} \\
\end{align}
\]

```{r logprey, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, e, d, K
    
    sliderInput("r_logprey", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    numericInput("K_logprey", label = "Carrying capacity of Prey", min = 1, value = 100),
    sliderInput("a_logprey", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("e_logprey", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 0.5),
    sliderInput("d_logprey", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
    
    
      
    ### Ask users for initial conditions -----
    numericInput("H_logprey", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_logprey", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----
    numericInput("t_logprey", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_logprey", label = "Display vector field?", 
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_logprey()), 
            renderPlot(np_logprey())
            # renderPlot(dabund_plot_logprey())
            )
)

# Set the initial population sizes
init_logprey <- reactive({
  c(H = input$H_logprey, P = input$P_logprey)
})

# Set the parameter values
pars_logprey <- reactive({
  c(r = input$r_logprey, K = input$K_logprey, 
    a = input$a_logprey, e = input$e_logprey, d = input$d_logprey)
  })

# Time over which to simulate model dynamics
time_logprey <- reactive({seq(0, input$t_logprey, by = .1)})

# Simulate the dynamics
out_logprey <- reactive({
    data.frame(ecoevoapps::run_predprey_model(
    time = time_logprey(),
    init = init_logprey(),
    params = pars_logprey()
  ))
})
  

# Reshape the data for plotting
out_long_logprey <- reactive({
  pivot_longer(out_logprey(), c(H,P), "Population")
})


# Plots ------

## make abundance thru time plot
plot_logprey <- reactive({
  ggplot(out_long_logprey()) + 
    geom_line(aes(x = time, y = value, color = Population), size = 2) + 
    scale_color_brewer(palette = "Set1") +
    ylab("Population size") 
})

## make isocline plot and add vector field depending on user input
vec_input_logprey <- reactive({
  vector_field_input(sim_df = out_logprey(), 
                     eq_func = logprey_eq, 
                     pars_for_eq_func = pars_logprey(),
                     vec_density = 15)
})

np_logprey <- reactive({
  
  if("Yes" %in% input$vectors_logprey) {
  vector_field(out_logprey(), vec_input_logprey()) +
    
    # the isoclines
    geom_abline(intercept = input$r_logprey/input$a_logprey, 
                slope = (-1 * input$r_logprey)/(input$a_logprey * input$K_logprey), 
                col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
    geom_vline(xintercept = input$d_logprey/(input$e_logprey*input$a_logprey), 
               col = brewer.pal(n = 3, name = "Set1")[2], size = 2) + 
    
    # the trace of the simulation and arrow for direction of the trace
    geom_path(aes(x = H, y = P), size = 2) +
    geom_segment(x = out_logprey()$H[min(5, round(length(time_logprey()))/50)], #did this min thing in case someone asks for very few timesteps
                 y = out_logprey()$P[min(5, round(length(time_logprey()))/50)],
                 xend = out_logprey()$H[min(5 + 1, round(length(time_logprey())/50) + 1)],
                 yend = out_logprey()$P[min(5 + 1, round(length(time_logprey())/50) + 1)],
                 arrow = arrow(length = unit(0.1, "npc")),
                 cex = 2) +
    
    # plot appearance
    xlab("Number of Prey") +
    ylab("Number of Predators") +
    coord_cartesian(xlim = c(min(out_logprey()$H), max(out_logprey()$H) + 1), 
                    ylim = c(min(out_logprey()$P), max(out_logprey()$P) + 1)) #need this line to show all vectors that go beyond plot limits
    
  } else {
    
    ggplot(out_logprey()) +
    # the isoclines
    geom_abline(intercept = input$r_logprey/input$a_logprey, 
                slope = (-1 * input$r_logprey)/(input$a_logprey * input$K_logprey), 
                col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
      geom_vline(xintercept = input$d_logprey/(input$e_logprey*input$a_logprey), 
                 col = brewer.pal(n = 3, name = "Set1")[2], size = 2) + 
    # the trace of the simulation and arrow for direction of the trace
    geom_path(aes(x = H, y = P), size = 2) +
    geom_segment(x = out_logprey()$H[min(5, round(length(time_logprey()))/50)], #did this min thing in case someone asks for very few timesteps
                 y = out_logprey()$P[min(5, round(length(time_logprey()))/50)],
                 xend = out_logprey()$H[min(5 + 1, round(length(time_logprey())/50) + 1)],
                 yend = out_logprey()$P[min(5 + 1, round(length(time_logprey())/50) + 1)],
                 arrow = arrow(length = unit(0.1, "npc")),
                 cex = 2) +
    
    # plot appearance
    xlab("Number of Prey") +
    ylab("Number of Predators") +
    coord_cartesian(xlim = c(min(out_logprey()$H), max(out_logprey()$H) + 1), 
                    ylim = c(min(out_logprey()$P), max(out_logprey()$P) + 1)) #need this line to show all vectors that go beyond plot limits
  }
  
})

```

### Type II Functional Response

This model presented here keeps the predator dynamics the same as in the L-V predator-prey model, but adds an additional layer of biological realism to the *predator*. Specifically, in the classic L-V model, the predator has a *Type 1 functional response*, which means that predator individuals never get satiated, and don't spend any time handling the prey. In this model, the predator has a **Type II functional response**, which means that the predator's efficiency in eating the prey is limited by the time it has to invest in processing the prey. 

\[
\begin{align}
\frac{dH}{dt} &= rH - \frac{aH}{1+aT_hH} P\\
\\
\frac{dP}{dt} &= eP \frac{aH}{1+aT_hH} - dP
\end{align}
\]

```{r params type ii, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$a$", 
               "$T_{h}$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Attack rate of the predator",
                 "Handling time of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a} +rT_hH\\
\\
H^* &= \frac{d}{ea - adT_h} \\
\end{align}
\]


```{r eruptions, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    sliderInput("r_lv_pred2", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .5),
    sliderInput("a_lv_pred2", label = "Predator attack rate", min = .001, max = 1.0, value = .1),
    sliderInput("T_h_lv_pred2", label = "Predator handling time", min = 0.001, max = 1.0, value = 0.2),
    sliderInput("e_lv_pred2", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 0.7),
    sliderInput("d_lv_pred2", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .3),
      
    ### Ask users for initial conditions -----
    numericInput("H_lv_pred2", label = "Initial population size of Prey", min = 1, value = 10),
    numericInput("P_lv_pred2", label = "Initial population size of Predator", min = 1, value = 10),
    
    ### Ask users for time to simulate ----    
    numericInput("t_lv_pred2", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_lv_pred2", label = "Display vector field?", 
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_lvpred2()),
            renderPlot(np_lvpred2())
  )
)

# Set the initial population sizes
init_lv_pred2 <- reactive({
  c(H = input$H_lv_pred2 , P = input$P_lv_pred2)
})

# Set the parameter values
pars_lv_pred2 <- reactive({
  c(r = input$r_lv_pred2, a = input$a_lv_pred2, 
    T_h = input$T_h_lv_pred2 , e = input$e_lv_pred2, 
    d = input$d_lv_pred2 )})

# Time over which to simulate model dynamics
time_lv_pred2 <- reactive({seq(0,input$t_lv_pred2,by = .1)})

# Simulate the dynamics
out_lv_pred2 <- reactive({
  data.frame(ecoevoapps::run_predprey_model(
    time = time_lv_pred2(),
    init = init_lv_pred2(),
    params = pars_lv_pred2()
  ))
})
  

# Reshape the data for plotting
out_long_lv_pred2 <- reactive({
  pivot_longer(out_lv_pred2(), c(H,P), "Population")
})

# Plots ------

## make abundance thru time plot
plot_lvpred2 <- reactive({
  ggplot(out_long_lv_pred2()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
    ylab("Population size")
})

## make isocline plot and add vector field depending on user input
vec_input_lv_pred2 <- reactive({
  vector_field_input(sim_df = out_lv_pred2(), 
                     eq_func = lv_pred2_eq, 
                     pars_for_eq_func = pars_lv_pred2())})

np_lvpred2 <- reactive({
  if("Yes" %in% input$vectors_lv_pred2) {
  vector_field(out_lv_pred2(), vec_input_lv_pred2()) +
    
    # the isoclines
    geom_abline(intercept = input$r_lv_pred2/input$a_lv_pred2, 
                slope = input$r_lv_pred2*input$T_h_lv_pred2, 
                col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
    geom_vline(xintercept = input$d_lv_pred2/(input$e_lv_pred2*input$a_lv_pred2 - input$a_lv_pred2*input$d_lv_pred2*input$T_h_lv_pred2), 
               col = brewer.pal(n = 3, name = "Set1")[2], size = 2) + 
    
    # the trace of the simulation and arrow for direction of the trace
    geom_path(aes(x = H, y = P), size = 2) +
    geom_segment(x = out_lv_pred2()$H[min(5, round(length(time_lv_pred2()))/50)], #did this min thing in case someone asks for very few timesteps
                 y = out_lv_pred2()$P[min(5, round(length(time_lv_pred2()))/50)],
                 xend = out_lv_pred2()$H[min(5 + 1, round(length(time_lv_pred2())/50) + 1)],
                 yend = out_lv_pred2()$P[min(5 + 1, round(length(time_lv_pred2())/50) + 1)],
                 arrow = arrow(length = unit(0.1, "npc")),
                 cex = 2) +
    
    # plot appearance
    xlab("Number of Prey") +
    ylab("Number of Predators") +
    coord_cartesian(xlim = c(min(out_lv_pred2()$H), max(out_lv_pred2()$H) + 1), ylim = c(min(out_lv_pred2()$P), max(out_lv_pred2()$P) + 1)) #need this line to show all vectors that go beyond plot limits
    
  } else {
    
    ggplot(out_lv_pred2()) +
    # the isoclines
    geom_abline(intercept = input$r_lv_pred2/input$a_lv_pred2, slope = input$r_lv_pred2*input$T_h_lv_pred2, col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
    geom_vline(xintercept = input$d_lv_pred2/(input$e_lv_pred2*input$a_lv_pred2 - input$a_lv_pred2*input$d_lv_pred2*input$T_h_lv_pred2), col = brewer.pal(n = 3, name = "Set1")[2], size = 2) + 
    
    # the trace of the simulation and arrow for direction of the trace
    geom_path(aes(x = H, y = P), size = 2) +
    geom_segment(x = out_lv_pred2()$H[min(5, round(length(time_lv_pred2()))/50)], #did this min thing in case someone asks for very few timesteps
                 y = out_lv_pred2()$P[min(5, round(length(time_lv_pred2()))/50)],
                 xend = out_lv_pred2()$H[min(5 + 1, round(length(time_lv_pred2())/50) + 1)],
                 yend = out_lv_pred2()$P[min(5 + 1, round(length(time_lv_pred2())/50) + 1)],
                 arrow = arrow(length = unit(0.1, "npc")),
                 cex = 2) +
    
    # plot appearance
    xlab("Number of Prey") +
    ylab("Number of Predators") +
    coord_cartesian(xlim = c(min(out_lv_pred2()$H), max(out_lv_pred2()$H) + 1), ylim = c(min(out_lv_pred2()$P), max(out_lv_pred2()$P) + 1)) #need this line to show all vectors that go beyond plot limits

  }
  
})

```

### Rosenzweig-MacArthur Model 

The Rosenzweig-MacArthur model combines both logistic growth for the prey as well as a Type II functional response for the predator.

\[
\begin{align}
\frac{dH}{dt} &= rH  \biggl(1-\frac{H}{K}\biggr) - \frac{aHP}{1+aT_hH} \\
\\
\frac{dP}{dt} &= e \frac{aHP}{1+aT_hH} - dP
\end{align}
\]

```{r params rosen-mac, echo=F}
pars_vars <- c("$H$", 
               "$P$", 
               "$r$", 
               "$K$", 
               "$a$", 
               "$T_{h}$", 
               "$e$", 
               "$d$")
descriptions <- c("Population size of the prey",
                 "Population size of the predator",
                 "Per capita growth rate of the prey",
                 "Carrying capacity of the prey",
                 "Attack rate of the predator",
                 "Handling time of the predator",
                 "Conversion efficiency of the predator",
                 "Death rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  

Equations for the zero net growth isoclines:
\[
\begin{align}
P^* &= \frac{r}{a}\bigg(1 - \frac{H}{K}\bigg)(1 + aT_hH)\\
\\
H^* &= \frac{d}{ea - adT_h} \\
\end{align}
\]


```{r rosenzweig-macarthur, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    ## r, a, T_h, e, d
    
    sliderInput("r_rm", label = "Per capita growth rate of Prey", min = .0001, max = 1.0, value = .2),
    numericInput("K_rm", label = "Prey carrying capacity", min = 1, value = 150),
    sliderInput("a_rm", label = "Predator attack rate", min = .001, max = 1.0, value = .02),
    sliderInput("T_h_rm", label = "Predator handling time", min = 0.001, max = 1.0, value = 0.3),
    sliderInput("e_rm", label = "Predator conversion efficiency", min = .001, max = 1.0, value = 0.6),
    sliderInput("d_rm", label = "Per capita death rate of Predator", min = .0001, max = 1.0, value = .4),
    
    
      
    ### Ask users for initial conditions -----
    #N1, N2
    numericInput("H_rm", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("P_rm", label = "Initial population size of Predator", min = 1, value = 25),
    
    ### Ask users for time to simulate ----
    numericInput("t_rm", label = "Timesteps", min = 1, value = 100),
    
    checkboxGroupInput("vectors_rm", label = "Display vector field?",
                       choices = c("Yes" = "Yes"), selected = "Yes")
    
  ),  
  
 
  mainPanel(renderPlot(plot_rm()),
            renderPlot(np_rm())
            )
  
)

# Set the initial population sizes
init_rm <- reactive({c(H = input$H_rm , P = input$P_rm)})

# Set the parameter values
pars_rm <- reactive({
  c(r = input$r_rm, K = input$K_rm, 
    a = input$a_rm, T_h = input$T_h_rm, 
    e = input$e_rm, d = input$d_rm )})

# Time over which to simulate model dynamics
time_rm <- reactive({seq(0,input$t_rm,by = .1)})

out_rm <- reactive({
    data.frame(ecoevoapps::run_predprey_model(
    time = time_rm(),
    init = init_rm(),
    params = pars_rm()
  ))
  })
  

# Reshape the data so that population sizes of both 
out_long_rm <- reactive({
  pivot_longer(out_rm(), c(H,P), "Population")
})


# Plots ------
## make abundance thru time plot
plot_rm <- reactive({
  ggplot(out_long_rm()) + 
  geom_line(aes(x = time, y = value, color = Population), size = 2) + 
  scale_color_brewer(palette = "Set1") +
    ylab("Population size") 
})

## make isocline plot and add vector field depending on user input
vec_input_rm <- reactive({
  vector_field_input(sim_df = out_rm(), eq_func = rm_predation_eq, pars_for_eq_func = pars_rm())
  })

np_rm <- reactive({
  
  if("Yes" %in% input$vectors_rm) {
  vector_field(out_rm(), vec_input_rm()) +
    
    # the isoclines
    stat_function(fun = function(x) (input$r_rm/input$a_rm)*(1 - x/input$K_rm)*(1 + input$a_rm * input$T_h_rm * x), col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
    geom_vline(xintercept = input$d_rm/(input$e_rm*input$a_rm - input$a_rm*input$d_rm*input$T_h_rm), col = brewer.pal(n = 3, name = "Set1")[2], size = 2) + 
    
    # the trace of the simulation and arrow for direction of the trace
    geom_path(aes(x = H, y = P), size = 2) +
    geom_segment(x = out_rm()$H[min(5, round(length(time_rm()))/50)], #did this min thing in case someone asks for very few timesteps
                 y = out_rm()$P[min(5, round(length(time_rm()))/50)],
                 xend = out_rm()$H[min(5 + 1, round(length(time_rm())/50) + 1)],
                 yend = out_rm()$P[min(5 + 1, round(length(time_rm())/50) + 1)],
                 arrow = arrow(length = unit(0.1, "npc")),
                 cex = 2) +
    
    # plot appearance
    xlab("Number of Prey") +
    ylab("Number of Predators") +
    coord_cartesian(xlim = c(min(out_rm()$H), max(out_rm()$H) + 1), ylim = c(min(out_rm()$P), max(out_rm()$P) + 1))  #need this line to show all vectors that go beyond plot limits

  
  } else {
    
    ggplot(out_rm()) +
    # the isoclines
    stat_function(fun = function(x) (input$r_rm/input$a_rm)*(1 - x/input$K_rm)*(1 + input$a_rm * input$T_h_rm * x), col = brewer.pal(n = 3, name = "Set1")[1], size = 2) +
    geom_vline(xintercept = input$d_rm/(input$e_rm*input$a_rm - input$a_rm*input$d_rm*input$T_h_rm), col = brewer.pal(n = 3, name = "Set1")[2], size = 2) + 
    
    # the trace of the simulation and arrow for direction of the trace
    geom_path(aes(x = H, y = P), size = 2) +
    geom_segment(x = out_rm()$H[min(5, round(length(time_rm()))/50)], #did this min thing in case someone asks for very few timesteps
                 y = out_rm()$P[min(5, round(length(time_rm()))/50)],
                 xend = out_rm()$H[min(5 + 1, round(length(time_rm())/50) + 1)],
                 yend = out_rm()$P[min(5 + 1, round(length(time_rm())/50) + 1)],
                 arrow = arrow(length = unit(0.1, "npc")),
                 cex = 2) +
    
    # plot appearance
    xlab("Number of Prey") +
    ylab("Number of Predators") +
    coord_cartesian(xlim = c(min(out_rm()$H), max(out_rm()$H) + 1), ylim = c(min(out_rm()$P), max(out_rm()$P) + 1)) #need this line to show all vectors that go beyond plot limits
  }
  
})

```

## Consumer Functional Response {.tabset}

Represents the relationship between consumer intake rate as a function of resource density

```{r functional response functions, echo=FALSE}
type1_fr <- function (H_init, H_final, a){
  to_return <- c(a*(H_init:H_final))
  return(to_return)
}

type2_fr <- function (H_init, H_final, a, Th){
  to_return <- c((a*(H_init:H_final)/(1+a*Th*(H_init:H_final))))
  return(to_return)
}

```

### Type I

Linear increase in intake rate with resource density, as shown by the equation:

$$f(H) = aH$$


```{r params type 1 fr, echo=F}
pars_vars <- c("$H$", 
               "$a$")
descriptions <- c("Population size of the prey",
                 "Attack rate of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  


```{r type 1 fr, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    # a
    sliderInput("a_fr_type1", label = "Predator attack rate", min = .001, max = 1.0, value = .02),
  
    ### Ask users for initial and final conditions -----
    #H_init, H_final
    
    numericInput("H_init_fr_type1", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("H_final_fr_type1", label = "Final population size of Prey", min = 1, value = 100), #Add warning: H_final>H_init 
    
  ),  
  
 mainPanel(renderPlot(plot_fr_type1())),

)


# Generate function output and data frame

out_fr_type1 <- reactive({type1_fr(H_init = input$H_init_fr_type1, 
                                   H_final = input$H_final_fr_type1,
                                   a = input$a_fr_type1)})

type1_fr_df <- reactive({data.frame(H = input$H_init_fr_type1:input$H_final_fr_type1,
                                    f_rate = out_fr_type1())})

#Plot

plot_fr_type1 <- reactive ({ggplot(type1_fr_df()) +
                            geom_line(aes(x = H, y=f_rate), size = 2) +
                            scale_color_brewer(palette = "Set1") +
                            xlab("Number of Prey") + 
                            ylab("Foraging Rate") +
                            ylim(0,50)})
   
```

### Type II

Decelerating intake rate, where the plateau represents consumer saturation:


$$g(H) = \frac {aH}{1+aT_hH}$$

```{r params type 2 fr, echo=F}
pars_vars <- c("$H$", 
               "$a$", 
               "$T_h$")
descriptions <- c("Population size of the prey",
                 "Attack rate of the predator",
                 "Handling time of the predator")
param_df <- data.frame(pars_vars, descriptions)
kable(x = param_df, format = "html", 
      col.names = c("Parameter/Variable", "Description")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
```  


```{r type 2 fr, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    ### Ask users for parameter values ----
    # a, Th
    
    sliderInput("a_fr_type2", label = "Predator attack rate", min = .001, max = 1.0, value = .02),
    sliderInput("T_h_fr_type2", label = "Predator handling time", min = 0.001, max = 1.0, value = 0.3),
  
    ### Ask users for initial and final conditions -----
    #H_init, H_final
    
    numericInput("H_init_fr_type2", label = "Initial population size of Prey", min = 1, value = 30),
    numericInput("H_final_fr_type2", label = "Final population size of Prey", min = 1, value = 100), #Add warning: H_final>H_init 
    
  ),  
  
 mainPanel(renderPlot(plot_fr_type2())),

)


# Generate function output and data frame

out_fr_type2 <- reactive({type2_fr(H_init = input$H_init_fr_type2, 
                                   H_final = input$H_final_fr_type2,
                                   a = input$a_fr_type2,
                                   Th = input$T_h_fr_type2)})

type2_fr_df <- reactive({data.frame(H = input$H_init_fr_type2:input$H_final_fr_type2,
                                    f_rate = out_fr_type2())})

#Plot

plot_fr_type2 <- reactive ({ggplot(type2_fr_df()) +
                            geom_line(aes(x = H, y=f_rate), size = 2) +
                            scale_color_brewer(palette = "Set1") +
                            xlab("Number of Prey") + 
                            ylab("Foraging Rate") +
                            ylim(0,4)})
   
```


## References 
- [Consumer Resource Dynamics](https://press.princeton.edu/books/paperback/9780691006574/consumer-resource-dynamics-mpb-36) by Murdoch, Briggs and Nisbet - digital copy may be available through a library.  
- [Graphical Representation and Stability Conditions of Predator-Prey Interactions](https://www.journals.uchicago.edu/doi/10.1086/282272), 1963, by M. L. Rosenzweig and R. H. MacArthur.   
- [Primer on functional response and numerical response](https://web.ma.utexas.edu/users/davis/375/popecol/lec10/funcresp.html)  
- [Some characteristics of simple types of predation and parasitism](https://hahana.soest.hawaii.edu/cmoreserver/summercourse/2010/documents/Holling_1959b.pdf), 1959, by C.S. Holling. 
