---
title: "Pulliam's Source-Sink Metapopulation Model"
author: "Xinyi Yan"
date: "June 28, 2020"
output: 
  html_document:
    css: bg.css
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
runtime: shiny
---
<div style="margin-bottom:10%;">
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
</div>

# Model overview
A metapopulation is a collection of spatially separated populations of a given species.  
The source-sink metapopulation model focuses on the "impact of **habitat-specific demographic rates** on population growth and regulation." (Pulliam 1988).   

In the simplified model scenario, the two populations differ in their fecundity due to different habitat quality. The **source** population has a high-quality habitat, in which it achieves high fecundity and overall positive population growth rate until it reaches a carrying capacity. The **sink** population has a low-quality habitat in which the population cannot sustain itself. 

The two populations are linked via emmigration/immigration, and their population growths are described in the next section. 

# Annual cycle & population growth
The population size at the **beginning of a year** is denoted $n$.  
During the summer breeding season, each adult produces $\beta$ offspring.  
Therefore, the **end-of-summer** population size is given by:
$$n_{\mathrm{end~of~summer}} = \mathrm{adults} + \mathrm{newborn~juveniles} = n + \beta n $$
  
The annual survival rate of adults is denoted $P_{A}$, and that of juveniles is denoted $P_{J}$. 

Therefore, the **end-of-year** population size is:
$$n_{\mathrm{end~of~year}} = \mathrm{survived~adults} + \mathrm{survived~juveniles} = P_{A}n + P_{J} \beta n $$
  
Thus, we can respectively write the end-of-year size of the source population (1) and sink population (2) as: 
$$n_{1\mathrm{~end~of~year}} = P_{A}n_{1} + P_{J} \beta_{1} n_{1} = \lambda_{1} n_{1}$$
$$n_{2\mathrm{~end~of~year}} = P_{A}n_{2} + P_{J} \beta_{2} n_{2} = \lambda_{2} n_{2}$$
where the $\lambda_i = (P_A + P_J\beta_i)$ represents the net annual growth rates of the respective population.  Populations can grow provided that $\lambda > 1$. 

Note that the Pulliam model assumes the source and sink sites differ only in their effect on the fecundity ($\beta$). The source population by definition has a positive population growth rate ($\lambda_1$ > 1) and the sink population by definition has a negative population growth rate ($\lambda$ < 1).     
  
If the source population exceeds the site carrying capacity ($N^{*}$), any excess individuals will **disperse** to the sink site, where there is no upper limit on population size (i.e. no carrying capacity). 
$$E_{\rm{from~sink}} = I_{\mathrm{to~source}} = n_{1\mathrm{~end~of~year}} - N^{*} $$
At the **beginning of the next year**, the source and sink population will become:  

- if source population has not reached carrying capacity:
$$ n_{1~\mathrm{next~year~initial}} =  P_{A}n_{1} + P_{J} \beta_{1} n_{1} = \lambda_{1} n_{1}$$
$$n_{2~\mathrm{next~year~initial}} = P_{A}n_{2} + P_{J} \beta_{2} n_{2} = \lambda_{2} n_{2}$$

- or, if the source population size exceeds the carrying capacity ($n_{1}$ = $N^*$):
$$ n_{1\mathrm{~next~year~initial}} = N^{*} $$
$$n_{2\mathrm{~next~year~initial}} = n_{2\mathrm{~end~of~year}} + I_{\mathrm{to~source}}= \lambda_{2} n_{2} + (\lambda_{1} n_{1} - N^{*}) = \lambda_{2} n_{2} + (\lambda_{1} - 1)N^{*}$$

From there, the annual cycle begins again.

## Table of model parameters

```{r paramtable, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)

paramTable <- data.frame(
  Parameter = c("$P_{a}$", "$P_{j}$", "$\\beta_{1}$", "$\\beta_{2}$", "$\\lambda_{1}$", "$\\lambda_{2}$", "$n_{1}$", "$n_{2}$", "$N_{1}$ or $N^{*}$"),
  
  Description = c("Probability of adults surviving winter", "Probability of juveniles surviving winter", "Fecundity at the source site", "Fecundity at the sink site", "Source population growth rate, given by $P_{a}$ + $\\beta_{1}$ $P_{j}$", "Sink population growth rate, given by $P_{a}$ + $\\beta_{2}$ $P_{j}$", "Initial source population size", "Initial sink population size", "Carrying capacity/equilibirum of source population"),
  
  Value.Range = c("0-1, constrained by $\\lambda$s", "0-1, constrained by $\\lambda$s", "> 0,  constrained by $\\lambda$s", "> 0, constrained by $\\lambda$s", "> 1", "< 1", "positive integers", "positive integers", "positive integers"))

kable(paramTable) %>%
  kable_styling(bootstrap_options = c("striped"))
```

# Simulation of the populations

## Interactive simulation plot
```{r, include=FALSE}
# Pulliam source-sink metapopulation model
library(tidyverse)
library(shiny) # need the (newest as of July 2020) version 1.5.0
```

```{r, include=FALSE}
# Write a function to calculate Nt+1 for both populations, using DISCRETE time
source_sink <- function(endtime,init,params) {
  with (as.list(c(endtime,init,params)), {
    # description of parameters/state variables:
    # pa = probability of adults surviving winter
    # pj = probability of juvenile surviving winter
    # beta1 = fecundity of population 1
    # beta2 = fecundity of population 2
    # n1 <- population of n1 (source by default)
    # N1 <- limiting breeding site for source population (equilibrium source population)
    # n2 <- population of n2 (sink by default)
    # assume sink population has unlimited breeding sites, equilibirum N2 will be calculated
    # when all breading sites are occupied, emigration happens
    
    n1 <- numeric(endtime)
    n2 <- numeric(endtime)
    e <- 0 # before breeding sites are all occupied at the source, emmigration is 0
    
    n1[1] <- n10
    n2[1] <- n20
    
    for (t in 2:endtime) {
      # at the start of the year, population = end of last year
      n10 <- n1[t-1]
      n20 <- n2[t-1]

      # growth cycle  of the year
      n1[t] <- n10*(pa + pj*beta1) 
      n2[t] <- n20*(pa + pj*beta2)
      # optional: rounding, keep populations integers 
      # n1[t] <- round( n1[t])
      # n2[t] <- round( n2[t])
      
      # migration cycle of the year
      if (n1[t] >= N1){
        
        e <- n1[t] - N1 #emmigration away from source, to sink
      }
      n1[t] <- n1[t] - e 
      n2[t] <- n2[t] + e
    }
    # return both n1 and n2
    return(list(n1, n2))
  })
}

```

```{r UIModules, echo=FALSE}
# use UI modeules for separate input for desktop vs mobile devices
# later each one will be under their own NameSpace
ssinputUI <- function(id) {
    tagList(
        ### ask for params (model parameters)
        ### params <- c(pa = .7, pj = .2, beta1 = 3, beta2 = 1, N1 = 300) 
        sliderInput(NS(id, "pa"), label = "Probability of adults surviving winter (Pa)", min = 0, max = 1, value = .7),
        sliderInput(NS(id, "pj"), label = "Probability of juveniles surviving winter (Pj)", min = 0, max = 1, value = .2),
        sliderInput(NS(id, "beta1"), label = "Fecundity at the source site (beta1)", min = 0, max = 100, value = 3),
        textOutput(NS(id, "suggest_beta1")), # suggested beta1 range given pa and pj
        sliderInput(NS(id, "beta2"), label = "Fecundity at the sink site (beta2)", min = 0, max = 50, value = 1),
        textOutput(NS(id, "suggest_beta2")), # suggested beta2 range given pa and pj
        numericInput(NS(id, "N1"), label = "maximum breeding sites at source", min = 1, value = 300),
        
        ### ask for init (inital conditions)
        ### init <- c(N = 20, P = 2)
        numericInput(NS(id, "n10"), label = "initial population at source (n1)", min = 0, value = 110),
        numericInput(NS(id, "n20"), label = "initial population at sink (n2)", min = 0, value = 100),
        
        ### ask for time (time to simulate)
        numericInput(NS(id, "endtime"), label = "years to simulate", min = 2, value = 50)
    ) # end of taglist (input)
}

# use UI modeules for separate simulation plots
sssimuUI <- function(id){
    tagList(
        plotOutput(NS(id, "simulation"))
    )
}

# use UI modeules for separate equilibirum output
ssequiUI <- function(id){
    tagList(
        actionButton(NS(id, "showAnswer"), "Show/update sink equilibirum", style = "color: rgb(201, 18, 18);"),
        textOutput(NS(id, "equilibrium"))
    ) # end of taglist
}
```

```{r serverModules, echo=FALSE}
# server modules for separate output based on respectve inputs (desktop vs mobile tabs)
ssServer <- function(id) {
    moduleServer(id, function(input, output, session) {
      
        ### first check if beta1 is valid for source population
        observeEvent(input$beta1, # check this input spot
          shinyFeedback::feedbackWarning( # gives warning if:
            "beta1", # this variable (beta1)
            input$pa + input$pj * input$beta1 <= 1, # if satisefy this condition
            "pa + pj * beta1 should >1 for source polulation, view suggestion.  
            Note that the violation does not stop the simulation. It just no longer satisefies the definition of a Pulliam source population" 
            # this warning message will show up
            # but simlulation will continue even with warning
            # if want simulation ONLY with valid input, use REQUIRE statement
          )  
        )
        ### suggested beta1 value based on pa and pj
        output$suggest_beta1 <- renderText(paste("Suggested input: beta1 >", round((1-input$pa) / input$pj, digits = 1)) )
        
        ### then check if beta2 is valid for sink population
        observeEvent(input$beta2,
          shinyFeedback::feedbackWarning(
            "beta2", 
            input$pa + input$pj * input$beta2 >= 1,
            "pa + pj * beta2 should <1 for sink polulation, view suggestion.  
            Note that the violation does not stop the simulation. It just no longer satisefies the definition of a Pulliam sink population"
          )  
        )
        ### suggested beta1 value based on pa and pj
        output$suggest_beta2 <- renderText(paste("Suggested input: beta2 <", round((1-input$pa) / input$pj, digits = 1)) )
        
        
        ### Set the initial population sizes
        #init <- c(n10 = 110, n20 = 100)
        #code tip: any time it's reactive, it has to be {} in ()
        init <- reactive({ c(n10 = input$n10, n20 = input$n20)}) 
        
        ### Set the parameter values
        #params <- c(pa = .7, pj = .2, beta1 = 3, beta2 = 1, N1 = 300) 
        params <- reactive({ c(pa = input$pa, pj = input$pj, 
                               beta1 = input$beta1, beta2 = input$beta2, N1 = input$N1) })
        ### Time over which to simulate model dynamics
        time <- reactive({ endtime = input$endtime })
        
        #code tip: now init and params are reactive, use them like functions "xx()"
        ss_df <- reactive({ data.frame(year = 1:time(), 
                                      source_sink(endtime = time(), init = init(), 
                                                    params = params())) })
        ss_out <- reactive({
           out <- ss_df()
           colnames(out) <- c("year", "source", "sink")
           out
         })
        
        ss_out_long <- reactive({ pivot_longer(ss_out(), c(source,sink), "population") })
        
        ### Plot
        ss_plot <- reactive({ ggplot(ss_out_long()) + 
          geom_line(aes(x = year, y = value, color = population), size = 2) + 
            ecoevoapps::theme_apps() + 
            scale_x_continuous(expand = c(0, 0, .1, 0)) +
            scale_y_continuous(expand = c(0, 0, .1, 0)) + 
            scale_color_brewer(palette = "Set1") +
            ylab("Population size")  })    
        
        output$simulation <- renderPlot(ss_plot(), height = 300)
        
        ### calculate equilibrium for sink population, display on click
        equi <- eventReactive(input$showAnswer, {round(-(input$pa + input$pj*input$beta1 - 1)*input$N1/(input$pa + input$pj*input$beta2 - 1), digits = 0)})
        output$equilibrium <- renderText({ paste(equi(), ". (Read section 4 for details.)") })
  })
}
```

```{r ssApp, echo=FALSE}
shinyApp( #to embed shiny app into the html document
  
    ### user interface part, ask for parameters and present the final outputs(suggested input, graph)
  ui = fluidPage(
    #withMathJax(), # not working so far -- for displaying special characters
    shinyFeedback::useShinyFeedback(), # feedback to ui
    
    # have two tabs for desktop vs mobile users, with different plot display format (fixed vs fluid)
    tabsetPanel(
      tabPanel("desktop full screen",
        fluidRow(
            column(4,
                wellPanel(width = "100%",
                ### ask for params (model parameters)
                # use UI module under NS "desktop"
                ssinputUI("desktop")
                )
            ), # end of left/input column
            
            column(8, 
                fixedPanel(right = 5, top = 30, width = "65%", sssimuUI("desktop") ) 
                # use UI module under NS "desktop" to give the plot
            ) # end of right/plot column
        ), # end of the first fluid row
        
        fixedRow(
          column(12, 
                 ### show calculated sink population equilibrium
                fixedPanel(right = 5, bottom = 15, width = "60%", height = 40,  
                  ssequiUI("desktop")
                ) # end of fixed panel
          ) # end of this column
        ) # end of the second fluid row
    ), # end of tab "desktop" 
    
    tabPanel("mobile devices",
        fluidRow(
          column(4,
              wellPanel(width = "100%",
              ### ask for params (model parameters)
              # use UI module under NS "mobile"
                ssinputUI("mobile")
              )
          ), # end of left/input column
          
          column(8, 
                 fluidRow(sssimuUI("mobile")),  # gives the plot
                 fluidRow(
                   wellPanel( 
                    ssequiUI("mobile")
                   ) # end of wellpanel
                 ) #end of equilibirium row
          ) # end of right/plot column
        )
      ) # end of tab "mobile"
    ) #end of tabset
  ), # end of ui

  ### server part, check validity of the input, generate output (graph) from the input
  server =  function(input, output, session) {
    # create modularized(separate) output from respective input for desktop vs mobile
    ssServer("desktop")
    ssServer("mobile")
  } # end of server
    
) # end of shiny app
```


# Population equilibirium

## Source population
Clearly, the equilibrium is the carrying capacity of the source site.
$$n_{1}^{*} = N^{*} $$

## Sink population
Recall that when the source population reaches equilibrium, the sink population annual increase can be written as:
$$n_{2~next~year~initial} =  (P_{A} + P_{J} \beta_{2}) n_{2} + (\lambda_{1} n_{1} - N^{*}) = \lambda_{2} n_{2} + (\lambda_{1} - 1)N^{*}$$
The above is a discrete time model, but if we consider a really long time, it becomes continuous. We represent the rate of change of sink population as the time deriative:
$$n_{2}^{'} =  (P_{A} + P_{J} \beta_{2} - 1) n_{2} + (\lambda_{1} n_{1} - N^{*}) = (\lambda_{2} - 1) n_{2} + (\lambda_{1} - 1)N^{*}$$
At equilibrium, the rate of change is 0 and we can solve for $n_{2}^{*}$:
$$0 = n_{2}^{'} = (\lambda_{2}-1) n_{2}^{*} + (\lambda_{1} - 1)N^{*} =  (P_{A} + P_{J} \beta_{2} - 1) n_{2}^{*} + (P_{A} + P_{J} \beta_{1} - 1)N^{*} $$
$$ n_{2}^{*} = -\frac{(P_{A} + P_{J} \beta_{1} - 1)N^{*}}{(P_{A} + P_{J} \beta_{2} - 1)}  $$
Calculate the equilibirum of your sink population based on your input and verify your answer using the simulation plot.  


# Reference & helpful resources:  
1. [Original Paper (Pulliam 1988)](https://www.journals.uchicago.edu/doi/abs/10.1086/284880)  
2. [Colorado State University Lecture Notes](https://sites.warnercnr.colostate.edu/gwhite/wp-content/uploads/sites/73/2017/06/Winkelman-Lecture-8.pdf)  
3. [UCLA Lecture Notes](https://faculty.eeb.ucla.edu/Sork/Env121Sork/SorkLec4_Metapop/Metapop.html#IV)


# DEV NOTES (will delete)

- require newest version of shiny package (1.5.0)

Possible additional features:   

ACTIVE/POSSIBLE:

- display greek letters (beta etc) inside shiny app

- make other metapoluation models and collect them together under tabs...

DONE:

- modularization of the simulation on different advice (due to fixed vs fluid layout) DONE

- currently the side bar style doesn't work due to adding the action button CANNOT, see next line for alternative

- ALTERNATIVE, use a different tab DONE

- model description DONE

- parameter table DONE

- click to show answer/ask for answer&check for equilibrium question. DONE, needs to fix position though  

- click to show equations (default collased)? NO NEED TO currently  

- Add some diagram (the one from the paper is great but copyright hmmm) or illustration (I have some drawings that's VAGUELY related to the topic) DONE  

- content list/tabs for the sections DONE  

