---
title: "Pulliam's Source-Sink Metapopulation Model"
author: "Xinyi Yan"
date: "June 28, 2020"
output: 
  html_document:
    css: bg.css
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
runtime: shiny
---
<div style="margin-bottom:10%;">
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
</div>

# Model overview
A metapopulation is a collection of spatially separated populations of the a given species.  
The source-sink metapopulation model focuses on the "impact of **habitat-specific demographic rates** on population growth and regulation." (Pulliam 1988).   
In the simplified model scenario, the two populations differ in their fecundity due to different habitat qualities. The **source** population has a high-quality habitat (but space limited), high fecundity, and overall positive population growth rate; vice versa for the **sink** population.      
The two populations are linked via emmigration/immigration, and their population growths are described in the next section. 

# Annual cycle & population growth
The population size at the **beinning of a year** is denoted by $n$.  
After summer (breeding season), each adult produces $\beta$ offspring.  
Therefore, the **end-of-summer** population size is:
$$n_{end~of~summer} = addults + newborn~juveniles = n + \beta n $$
  
The adults have an average annual survival rate of $P_{A}$, and juveniles of $P_{J}$. 

Therefore, the **end-of-year** population size is:
$$n_{end~of~year} = survived~addults + survived~juveniles = P_{A}n + P_{J} \beta n $$
  
Thus, we can respectively write the end-of-year source population (1) and sink population (2) as: 
$$n_{1~end~of~year} = P_{A}n_{1} + P_{J} \beta_{1} n_{1} = \lambda_{1} n_{1}$$
$$n_{2~end~of~year} = P_{A}n_{2} + P_{J} \beta_{2} n_{2} = \lambda_{2} n_{2}$$
where the $\lambda s$ are the annual growth rates of the respective population.  
Note that the Pulliam model assumes the source and sink sites differ only in their effect on the fecundity ($\beta$); the source population has a positive population growht rate ( $\lambda$ > 1) and the sink population has a negative population growth rate ($\lambda$ < 1).     
  
If the source population exceeds the site carrying capacity, any excess indivuduals at the source site (limited high-quality habitat, carrying capacity $N^{*}$) will **disperse** to the sink site (unlimited carrying capacity). 
$$E_{from~sink} = I_{to~source} = n_{1~end~of~year} - N^{*} $$
At the **beginning of the next year**, the source and sink population will become:  

- if source population has not reached carrying capacity:
$$ n_{1~next~year~initial} =  P_{A}n_{1} + P_{J} \beta_{1} n_{1} = \lambda_{1} n_{1}$$
$$n_{2~next~year~initial} = P_{A}n_{2} + P_{J} \beta_{2} n_{2} = \lambda_{2} n_{2}$$

- or, if having reached carrying capacity ($n_{1}$ = $N^*$):
$$ n_{1~next~year~initial} = N^{*} $$
$$n_{2~next~year~initial} = n_{2~end~of~year} + I_{to~source}= \lambda_{2} n_{2} + (\lambda_{1} n_{1} - N^{*}) = \lambda_{2} n_{2} + (\lambda_{1} - 1)N^{*}$$

From there, the annual cycle begins again.

# Simulation of the populations
```{r, include=FALSE}
# Pulliam source-sink metapopulation model
library(tidyverse)
library(shiny)

# Write a function to calculate Nt+1 for both populations, using DISCRETE time
source_sink <- function(endtime,init,params) {
  with (as.list(c(endtime,init,params)), {
    # description of parameters/state variables:
    # pa = probability of adults surviving winter
    # pj = probability of juvenile surviving winter
    # beta1 = fecundity of population 1
    # beta2 = fecundity of population 2
    # n1 <- population of n1 (source by default)
    # N1 <- limiting breeding site for source population (equilibrium source population)
    # n2 <- population of n2 (sink by default)
    # assume sink population has unlimited breeding sites, equilibirum N2 will be calculated
    # when all breading sites are occupied, emigration happens
    
    n1 <- numeric(endtime)
    n2 <- numeric(endtime)
    e <- 0 # before breeding sites are all occupied at the source, emmigration is 0
    
    n1[1] <- n10
    n2[1] <- n20
    
    for (t in 2:endtime) {
      # at the start of the year, population = end of last year
      n10 <- n1[t-1]
      n20 <- n2[t-1]

      # growth cycle  of the year
      n1[t] <- n10*(pa + pj*beta1) 
      n2[t] <- n20*(pa + pj*beta2)
      # optional: rounding, keep populations integers 
      # n1[t] <- round( n1[t])
      # n2[t] <- round( n2[t])
      
      # migration cycle of the year
      if (n1[t] >= N1){
        
        e <- n1[t] - N1 #emmigration away from source, to sink
      }
      n1[t] <- n1[t] - e 
      n2[t] <- n2[t] + e
    }
    # return both n1 and n2
    return(list(n1, n2))
  })
}

```

```{r, echo=FALSE}
# ui (with feedback), & plotting
shinyApp( #to embed shiny app into the html document
  
    ### user interface part, ask for parameters and present the final outputs(suggested input, graph)
    ui = fluidPage(
    #withMathJax(), # not working so far -- for displaying special characters
    shinyFeedback::useShinyFeedback(), # feedback to ui
    #div(titlePanel("Simulation: Pulliam's Source/Sink Metapopulation Dynamic", style = "position:fixed;width:inherit;")),
    #fixedPanel(height = 50, width = "100%","Simulation: Pulliam's Source/Sink Metapopulation Dynamic"),
    
    sidebarLayout(
      sidebarPanel(
        # style of the side panel
        style = "float:left;position:flex;width = 10%;",
        
        ### ask for params (model parameters)
        ### params <- c(pa = .7, pj = .2, beta1 = 3, beta2 = 1, N1 = 300) 
        sliderInput("pa", label = "Probability of adults surviving winter (Pa)", min = 0, max = 1, value = .7),
        sliderInput("pj", label = "Probability of juveniles surviving winter (Pj)", min = 0, max = 1, value = .2),
        sliderInput("beta1", label = "Fecundity at the source site (beta1)", min = 0, max = 100, value = 3),
        textOutput("suggest_beta1"),
        sliderInput("beta2", label = "Fecundity at the sink site (beta2)", min = 0, max = 50, value = 1),
        textOutput("suggest_beta2"),
        numericInput("N1", label = "maximum breeding sites at source", min = 1, value = 300),
        
        ### ask for init (inital conditions)
        ### init <- c(N = 20, P = 2)
        numericInput("n10",label = "initial population at source", min = 0, value = 110),
        numericInput("n20",label = "initial population at sink", min = 0, value = 100),
        
        ### ask for time (time to simulate)
        numericInput("endtime", label = "years to simulate", min = 2, value = 50),
        
        ### show calculated sink population equilibrium
        actionButton("showAnswer", "Display/update sink equilibirum", style = "color: blue;"),
        textOutput("equilibrium")
      ),
    
      ### Show a plot of the simulated population
      mainPanel(
        style = "margin-left:35%;position:fixed;width:65%",
        plotOutput("simulation") # gives the plot
        #the output name "simulation" is a renderPlot object desined in the server part
      )
      
      # fixedPanel(
      #   # #h3("simulation plot", align = "center"),
      #   # right = 10, width = "60%", plotOutput("simulation"))
     
    )
  ), # end of ui

  ### server part, check validity of the input, generate output (graph) from the input
  server = function(input, output, session) {
    ### first check if beta1 is valid for source population
    observeEvent(input$beta1, # check this input spot
      shinyFeedback::feedbackWarning( # gives warning if:
        "beta1", # this variable (beta1)
        input$pa + input$pj * input$beta1 <= 1, # satisefy this condition
        "pa + pj * beta1 should >1 for source polulation, view suggestion.  
        Note that the violation does not stop the simulation. It just no longer satisefies the definition of a Pulliam source population" 
        # this warning message will show up
        # but simlulation will continue even with warning
        # if want simulation ONLY with valid input, use REQUIRE statement
      )  
    )
    output$suggest_beta1 <- renderText(paste("Suggested input: beta1 >", round((1-input$pa) / input$pj, digits = 1)) )
    
    ### then check if beta2 is valid for sink population
    observeEvent(input$beta2,
      shinyFeedback::feedbackWarning(
        "beta2", 
        input$pa + input$pj * input$beta2 >= 1,
        "pa + pj * beta2 should <1 for sink polulation, view suggestion.  
        Note that the violation does not stop the simulation. It just no longer satisefies the definition of a Pulliam sink population"
      )  
    )
    output$suggest_beta2 <- renderText(paste("Suggested input: beta2 <", round((1-input$pa) / input$pj, digits = 1)) )
    
    ### Set the initial population sizes
    #init <- c(n10 = 110, n20 = 100)
    #any time it's reactive, it has to be {} in ()
    init <- reactive({ c(n10 = input$n10, n20 = input$n20)}) 
    
    ### Set the parameter values
    #params <- c(pa = .7, pj = .2, beta1 = 3, beta2 = 1, N1 = 300) 
    params <- reactive({ c(pa = input$pa, pj = input$pj, 
                           beta1 = input$beta1, beta2 = input$beta2, N1 = input$N1) })
    ### Time over which to simulate model dynamics
    time <- reactive({ endtime = input$endtime })
    
    #now init and params are reactive, use them like functions "xx()"
    ss_df <- reactive({ data.frame(year = 1:time(), 
                                  source_sink(endtime = time(), init = init(), 
                                                params = params())) })
    ss_out <- reactive({
       out <- ss_df()
       colnames(out) <- c("year", "source", "sink")
       out
     })
    
    ss_out_long <- reactive({ pivot_longer(ss_out(), c(source,sink), "population") })
    
    # Plot
    ss_plot <- reactive({ ggplot(ss_out_long()) + 
      geom_line(aes(x = year, y = value, color = population), size = 2) + 
      scale_color_brewer(palette = "Set1") +
      ylab("Population size")  })    
    
    output$simulation <- renderPlot(ss_plot())
    
    ### calculate equilibrium for sink population, display on click
    equi <- eventReactive(input$showAnswer, {round(-(input$pa + input$pj*input$beta1 - 1)*input$N1/(input$pa + input$pj*input$beta2 - 1), digits = 0)})
    output$equilibrium <- renderText({ paste(equi(), ". (Read the next section for details.)") })
    
  } # end of server
    
)
```

# Population equilibirium

## Source population
Clearly, the equilibrium is the carrying capacity of the source site.
$$n_{1}^{*} = N^{*} $$

## Sink population
Recall that when the source population reaches equilibrium, the sink population annual increase can be written as:
$$n_{2~next~year~initial} =  (P_{A} + P_{J} \beta_{2}) n_{2} + (\lambda_{1} n_{1} - N^{*}) = \lambda_{2} n_{2} + (\lambda_{1} - 1)N^{*}$$
The above is a discrete time model, but if we consider a really long time, it becomes continuous. We represent the rate of change of sink population as the time deriative:
$$n_{2}^{'} =  (P_{A} + P_{J} \beta_{2} - 1) n_{2} + (\lambda_{1} n_{1} - N^{*}) = (\lambda_{2} - 1) n_{2} + (\lambda_{1} - 1)N^{*}$$
At equilibrium, the rate of change is 0 and we can solve for $n_{2}^{*}$:
$$0 = n_{2}^{'} = (\lambda_{2}-1) n_{2}^{*} + (\lambda_{1} - 1)N^{*} =  (P_{A} + P_{J} \beta_{2} - 1) n_{2}^{*} + (P_{A} + P_{J} \beta_{1} - 1)N^{*} $$
$$ n_{2}^{*} = -\frac{(P_{A} + P_{J} \beta_{1} - 1)N^{*}}{(P_{A} + P_{J} \beta_{2} - 1)}  $$
Calculate the equilibirum of your sink population based on your input and verify your answer using the simulation plot.  


# Reference & helpful resources:  
1. [Original Paper (Pulliam 1988)](https://www.journals.uchicago.edu/doi/abs/10.1086/284880)  
2. [Colorado State University Lecture Notes](https://sites.warnercnr.colostate.edu/gwhite/wp-content/uploads/sites/73/2017/06/Winkelman-Lecture-8.pdf)  
3. [UCLA Lecture Notes](https://faculty.eeb.ucla.edu/Sork/Env121Sork/SorkLec4_Metapop/Metapop.html#IV)


# DEV NOTES
Possible additional features:   

- currently the side bar style doesn't work due to adding the action button

- model description DONE

- parameter table

- click to show answer/ask for answer&check for equilibrium question. DONE, needs to fix position though  

- click to show equations (default collased)?  

- Add some diagram (the one from the paper is great but copyright hmmm) or illustration (I have some drawings that's VAGUELY related to the topic) DONE  

- content list/tabs for the sections DONE  

- or make other metapoluation models and collect them together under tabs...
