---
title: "Structured Populations"
output: html_document
runtime: shiny
---

```{r, include = F}

# Set default parameters for R chunks in Rmd
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

# Required packages and functions
library(diagram) # plot diagrams
# library(gridExtra) # arrange outpupt plots into a grid
library(ggplot2) # produce nice graphics
library(ecoevoapps)
library(patchwork)
# library(tidyr)
# library(dplyr)
# library(reshape2) # help with matrix to data frame conversion 
# library(ggplotify) # convert base plots into graphic objects
# library(cowplot) # arrange plots into a grid
# source("struct_pop_function.R")

```

## Intro
Sometimes populations are best modeled as a collection of age or size classes ([Tenhumberg, 2010](https://www.nature.com/scitable/knowledge/library/ignoring-population-structure-can-lead-to-erroneous-15128671/)). The growth of such structured populations depends on the transition rates from one class to another. By wiggling the values of these rates we can start finding what are the transitions that impact the most the growth of the whole population (sensitivity analysis).

### The Leslie matrix
This is a convenient way of organizing age transition rates, plus it has interesting properties: 1) to simulate what the population strucure will look like one step ahead in time, one simply needs to  multiply the Leslie matrix by the matrix with the initial population sizes; 2) the largest eigenvalue of the Leslie matrix is $\lambda$, the asymptotic growth rate for the whole population ($\lambda<1$ it declines, $\lambda>1$ it grows, $\lambda=1$ it stays the same); 3) the eigenvector that corresponds to the largest eigenvalue gives the stable population structure, i.e., the proportions of each age class after the population reached equilibrium.

```{r}

### APP
sidebarLayout(
  ## INPUT
  sidebarPanel(
      
    # Initial sizes
    numericInput(inputId = "N1",
                  label = "Initial number of individuals at age 1",
                  min = 0,
                  max = NA,
                  value = 100),
    numericInput(inputId = "N2",
                   label = "Initial number of individuals at age 2",
                   min = 0,
                   max = NA,
                   value = 100),
    numericInput(inputId = "N3",
                   label = "Initial number of individuals at age 3",
                   min = 0,
                   max = NA,
                   value = 100),
      
    # Leslie matrix
    numericInput(inputId = "F1",
                  label = "Fecundity at age 1",
                  min = 0,
                  max = NA,
                  value = 0),
    numericInput(inputId = "F2",
                  label = "Fecundity at age 2",
                  min = 0,
                  max = NA,
                  value = 8),          
    numericInput(inputId = "F3",
                  label = "Fecundity at age 3",
                  min = 0,
                  max = NA,
                  value = 1),
    sliderInput(inputId = "S12",
                  label = "Survival from age 1 to 2 (%)",
                  min = 1,
                  max = 100,
                  value = 40),
    sliderInput(inputId = "S23",
                  label = "Survival from age 2 to 3 (%)",
                  min = 0,
                  max = 100,
                  value = 80),
    width=4),
  
  ## OUTPUT
  mainPanel(
    # renderPlot({ diagr() }, width=500),
    
    p("Leslie matrix"), 
    renderTable({ Leslie_matrix()}),
    p("_____________________"),
    
    p("Asymptotic population growth rate"),
    renderText({ eigenval() }),
    p("_____________________"), 
    
    p("Stable population structure"),
    renderTable({ eigenvec() }),
    
    renderPlot({ plots() }),
    
    width=8
    )
  
  )


## Leslie matrix
Leslie_matrix <- reactive({
  matrix(
      c(input$F1     ,      input$F2, input$F3,
        input$S12/100,             0,        0,
                    0, input$S23/100,        0),
      ncol = 3, byrow = T
      )
  })
init <- reactive({
  c(input$N1,input$N2,input$N3)
})

## Largest eigenvalue
eigenval <- reactive({
  paste0("Eigenvalue = ",round(eigen(Leslie_matrix())$values,2)[1])
  })

## Eigenvector corresponding to largest eigenvalue
tmp <- reactive({
  round(eigen(Leslie_matrix())$vectors[,1]*100/
          sum(eigen(Leslie_matrix())$vectors[,1]))
})

eigenvec <- reactive({
  matrix(
    c("Eigenvector", round(eigen(Leslie_matrix())$vectors[,1],2),
      "%", tmp()[1], tmp()[2], tmp()[3]),
    ncol=4, byrow=T,
    dimnames = list(NULL, c(" ","Age 1","Age 2", "Age 3"))
  )
})

## Plots
# Life cycle diagram
# diagr <- reactive({
#   plotmat(A = Leslie_matrix(), pos = 3, curve = 0.5, lwd = 1.5, my = -0.1,
#              name = c(expression(Age[1]),
#                       expression(Age[2]),
#                       expression(Age[3])), 
#              arr.len = 0.2, arr.width = 0.25, arr.lwd = 2, 
#              arr.type = "simple", self.lwd = 2, self.shiftx = 0.115, 
#              self.shifty = 0.1, self.cex = .5, box.size = 0.1, 
#              dtext = 0.2, box.lwd = 3.5, main="Life cycle diagram"
#              )
#   })

# Simulation plots

pop_trajectory <- reactive({
  ecoevoapps::run_structured_population_simulation(leslie_mat = Leslie_matrix(), init = init(), time = 1000)
})

plot_pop_trajectory <- reactive({
  ecoevoapps::plot_structured_population_size(pop_trajectory())
})

plot_pop_agedist <- reactive({
  ecoevoapps::plot_structured_population_agedist(pop_trajectory(), leslie_mat = Leslie_matrix())
})

plot_pop_lambda <- reactive({
  plot_structured_population_lambda(pop_trajectory(), leslie_mat = Leslie_matrix())
})

plots <- reactive({
  plot_pop_trajectory() + plot_pop_agedist() + plot_pop_lambda() + guide_area() +
    plot_layout(guides = "collect")
})

```


--------------

App maintained by Marcel Vaz (marcelcvaz@gmail.com)
