---
title: "Discrete population growth models"
output: html_document
runtime: shiny
---

***[中文](https://ecoevoapps.shinyapps.io/population_growth_discrete_ch/) | [Español](https://ecoevoapps.shinyapps.io/population_growth_discrete_es/) | English | [português](https://ecoevoapps.shinyapps.io/population_growth_discrete_pt) | [Turkish](https://ecoevoapps.shinyapps.io/population_growth_discrete_tr/) ***


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(ggplot2)
library(deSolve)
library(ecoevoapps)
library(patchwork)
library(latex2exp)
theme_set(ecoevoapps::theme_apps())
```

## Discrete exponential growth 

$N_{t+1} = \lambda N_t$

```{r exponential-logistic, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    # User defined lambda -------
    sliderInput("lambda", label = "Population growth rate (lambda):",
                min = 0.01, max = 2, value = 1.2, step = .01),
    # User defined N0 -------
    numericInput("N0", label = "Initial population size (N0)",
                min = 1, max = 100, value = 1),
    # User defined time ---------
    numericInput("time", label = "Number of time steps to project",
                min = 1, max = 1000, value = 30)
    ),

  # Panel of plots -----
  mainPanel(
    renderPlot({plots_to_print()}, width = 450, height = 450)
  )
)

# Generate trajectories for sp 1 --------

exponential_pop_df <- reactive({
  run_discrete_exponential_model(N0 = input$N0, lambda = input$lambda, time = input$time)
})


# Make plot -----------
trajaectory_plot <- reactive({
  plot_discrete_population_growth(exponential_pop_df()) +
    ylab(latex2exp::TeX("Population size at time $t$ ($N_t$)")) +
    labs(caption = latex2exp::TeX(paste0("Parameter values: $\\lambda = $", input$lambda)),
         title = "Population growth with the discrete exponential model")
})

# Make a list of plots  ----
plots_to_print <- reactive({trajaectory_plot()})

```

## Discrete logistic growth  {.tabset}

Next, we can explore models of population growth where the growth rate of a population depends on the current population size. There are several different canonical models that track the dynamics of populations whose growth rates decrease as population size increases. 

### Standard discrete logistic model

One way to model logistic growth in a population with discrete growth is with the [logistic map](https://en.wikipedia.org/wiki/Logistic_map), which is a discrete time analog of the continuous logistic equation:  

$$N_{t+1} = r_dN_t\left(1-\frac{N_t}{K}\right)$$

This formulation was notably studied by Robert May in the classic 1976 paper "[Simple mathematical models with very complicated dynamics](http://abel.harvard.edu/archive/118r_spring_05/docs/may.pdf)", in which he showed the potential for this model to generate [chaotic](https://en.wikipedia.org/wiki/Logistic_map#Chaos_and_the_logistic_map) dynamics.

```{r discrete-logistic, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    # User defined lambda -------
    sliderInput("rd_dl", label = "Population growth rate (rd):",
                min = 0.01, max = 4.5, value = 1.05, step = .01),
    # User defined N0 -------
    numericInput("N0_dl", label = "Initial population size (N0)",
                min = 1, max = 100, value = 1),
    # User defined carrying capacity (K) ----------    
    numericInput("K_dl", label = "Carrying capacity (K)",
                 min = 1, max = 1000, value = 100),
    # User defined time ---------
    numericInput("time_dl", label = "Number of time steps to project",
                min = 1, max = 1000, value = 30)

    ),

  # Panel of plots -----
  mainPanel(
    renderPlot({plots_to_print_dl()}, width = 450, height = 800)
  )
)

# Get user defined parameters for discrete logistic  ------

params_dl <- reactive({c(rd = input$rd_dl, K = input$K_dl)})
dl_df <- reactive({
  run_discrete_logistic_model(N0 = input$N0_dl,
                                       params = params_dl(),
                                       time = input$time_dl)


})
# Generate trajectories for discrete logistic --------
trajaectory_plot_dl <- reactive({
  plot_discrete_population_growth(dl_df()) +
    labs(title = "Population growth with the discrete logistic model")
})

# cobweb for standard discrete logistic model -----
cobweb_dl <- reactive({
  plot_discrete_population_cobweb(dl_df(),
                                  params_vec = params_dl(),
                                  model_type = "discrete_logistic")
})

# Make plot caption ----
plot_caption_dl <- reactive({
    latex2exp::TeX(paste0("Parameter values: $r_d = $", input$rd_dl, "; K = ", input$K_dl))
})
# Make a list of plots  ----
plots_to_print_dl <- reactive({{trajaectory_plot_dl()/cobweb_dl()} + 
    labs(caption = plot_caption_dl())})
```

### Ricker model

Another way to model a population with discrete growth with a carrying capacity is with the Ricker model, originally described in a [1954 paper](https://www.nrcresearchpress.com/doi/abs/10.1139/f54-039) about stock and recruitment in fisheries.  



$$N_{t+1} = N_t e^{(r (1 - N_t/K))}$$ 

```{r ricker-model, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    # User defined lambda -------
    sliderInput("r_ricker", label = "Population growth rate (r):",
                min = 0.01, max = 3, value = 1.01, step = .01),
    # User defined N0 -------
    numericInput("N0_ricker", label = "Initial population size (N0)",
                min = 1, max = 100, value = 1),
    # User defined carrying capacity (K) ----------    
    numericInput("K_ricker", label = "Carrying capacity (K)",
                 min = 1, max = 1000, value = 100),
    # User defined time ---------
    numericInput("time_ricker", label = "Number of time steps to project",
                min = 1, max = 1000, value = 30, step = 5)

    ),

  # Panel of plots -----
  mainPanel(
    renderPlot({plots_to_print_ricker()}, width = 450, height = 800)
  )
)
params_ricker <- reactive({c(rd = input$r_ricker, K = input$K_ricker)})

ricker_df <- reactive({
  run_ricker_model(N0 = input$N0_ricker,
                                 params = params_ricker(),
                                 time = input$time_ricker)
})
# Generate trajectories for discrete logistic --------
trajaectory_plot_ricker <- reactive({
  plot_discrete_population_growth(ricker_df()) +
    labs(title = "Population growth with the Ricker model")
})

# cobweb plot for ricker model -----
cobweb_ricker <- reactive({
    plot_discrete_population_cobweb(ricker_df(),
                                  params_vec = params_ricker(),
                                  model_type = "ricker")
})

# Make plot caption ----
plot_caption_ricker <- reactive({
    latex2exp::TeX(paste0("Parameter values: $r = $", input$r_ricker, "; K = ", input$K_ricker))
})

# Make a list of plots  ----

plots_to_print_ricker <- reactive({{
  trajaectory_plot_ricker()/cobweb_ricker()} + 
    labs(caption = plot_caption_ricker())})

```

### Beverton Holt model

A third classic model for population growth in discrete time is the Beverton-Holt model, which was also developed in the context of fisheries management:

$$N_{t+1} = \frac{RN_t}{1+\left(\frac{R-1}{K}\right)N_t}$$

```{r beverton-holt, echo=FALSE}
sidebarLayout(
  sidebarPanel(
    # User defined lambda -------
    sliderInput("r_bh", label = "Population growth rate (R):",
                min = 0.01, max = 3, value = 1.25, step = .01),
    # User defined N0 -------
    numericInput("N0_bh", label = "Initial population size (N0)",
                min = 1, max = 100, value = 1),
    # User defined carrying capacity (K) ----------    
    numericInput("K_bh", label = "Carrying capacity (K)",
                 min = 1, max = 1000, value = 100),
    # User defined time ---------
    numericInput("time_bh", label = "Number of time steps to project",
                min = 1, max = 1000, value = 30, step = 5)

    ),

  # Panel of plots -----
  mainPanel(
    renderPlot({plots_to_print_bh()}, width = 450, height = 800)
  )
)
# Get user defined parameters for discrete logistic  ------
params_bh <- reactive({c(rd = input$r_bh, K = input$K_bh)})
bh_df <- reactive({
  run_beverton_holt_model(N0 = input$N0_bh, 
                          params = params_bh(),
                          time = input$time_bh)
})
# Generate trajectories for beverton-holt model --------
trajaectory_plot_bh <- reactive({
  plot_discrete_population_growth(bh_df()) +
    labs(title = "Population growth with the Beverton-Holt model")
})

# Generate cobweb for beverton-holt model --------
cobweb_bh <- reactive({
  plot_discrete_population_cobweb(bh_df(),
                                  params_vec = c(rd = input$r_bh, K = input$K_bh),
                                  model_type = "beverton_holt") 
})

# Make plot caption ----
plot_caption_bh <- reactive({
    latex2exp::TeX(paste0("Parameter values: $r = $", input$r_bh, "; K = ", input$K_bh))
})

# Make a list of plots  ----

plots_to_print_bh <- reactive({{trajaectory_plot_bh()/cobweb_bh()} + 
    labs(caption = plot_caption_bh())})

```

## References

"[Simple mathematical models with very complicated dynamics](http://abel.harvard.edu/archive/118r_spring_05/docs/may.pdf)", Robert May, 1976.

"[Stock and Recruitment](https://www.nrcresearchpress.com/doi/abs/10.1139/f54-039)", Bill Ricker, 1954.

[Course notes on the Ricker model](http://courses.ecology.uga.edu/ecol4000-fall2018/wp-content/uploads/sites/22/2018/08/Chapter-3-complex-dynamics.pdf) from UGA course on population ecology 

Dr. Sebastian Bonhoeffer's [lecture notes](https://ethz.ch/content/dam/ethz/special-interest/usys/ibz/theoreticalbiology/education/learningmaterials/701-1424-00L/lecture-script-eep_2011.pdf) on Ecology and Evolution (see Ch. 1) 

Dr. Jan Engelstädter's [e-book on analysis of biological data](https://bookdown.org/janengelstaedter/biol3360modelling3/discrete-time-models-in-one-variable.html). 

-------------


```{r, results='asis', echo = F}
suppressWarnings(ecoevoapps::print_app_footer())
```


