---
title: "Mutualism"
output: html_document
runtime: shiny
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Mutualism is a major type of species interaction in which two or more interacting species confer a net beneficial effect on one another. Mutualisms can occur in various forms, including (but not limited to) plant-pollinator interactions, dispersal of seeds by animal vectors, and defense from predators. For example, fig wasps pollinate, lay eggs in, and feed on figs ([Cook & Rasplus 2003](https://doi.org/10.1016/S0169-5347(03)00062-4)). Another example is the mutualism between ants and acacia plants, in which ants feed on and shelter in acacias while protecting them from other herbivorous insects ([Janzen 1966](https://doi.org/10.2307/2406628)). A detailed overview of mutualisms can be found in [Bronstein (2015)](https://doi.org/10.1093/acprof:oso/9780199675654.001.0001).

This app explores a phenomenological (i.e. no explicit mechanism) model of direct mutualism between two species.

```{r app}
##### UI #####

# Input panel
wellPanel(
  fluidRow(
    column(4, numericInput(inputId = "r1",
                           label = HTML("r<sub>1</sub>: Per capita growth rate of sp. 1"),
                           value = 0.5,
                           step = 0.1)),
    column(4, numericInput(inputId = "r2",
                           label = HTML("r<sub>2</sub>: Per capita growth rate of sp. 2"),
                           value = 0.5,
                           step = 0.1)),
    column(4, numericInput(inputId = "N1",
                           label = "Initial population size of sp. 1",
                           value = 20,
                           min = 0,
                           step = 1))),
  fluidRow(
    column(4, numericInput(inputId = "a12",
                           label = HTML("&alpha;<sub>12</sub>: Effect of sp. 2 on sp. 1"),
                           value = 0.8,
                           min = 0,
                           step = 0.1)),
    column(4, numericInput(inputId = "a21",
                           label = HTML("&alpha;<sub>21</sub>: Effect of sp. 1 on sp. 2"),
                           value = 0.4,
                           min = 0,
                           step = 0.1)),
    column(4, numericInput(inputId = "N2",
                           label = "Initial population size of sp. 2",
                           value = 40,
                           min = 0,
                           step = 1))),
  fluidRow(
    column(4, numericInput(inputId = "d1",
                           label = HTML("d<sub>1</sub>: Rate of self-limitation for sp. 1"),
                           value = 0.02,
                           min = 0,
                           step = 0.01)),
    column(4, numericInput(inputId = "d2",
                           label = HTML("d<sub>2</sub>: Rate of self-limitation for sp. 2"),
                           value = 0.01,
                           min = 0,
                           step = 0.01)),
    column(4, numericInput(inputId = "time",
                           label = "Length of simulation",
                           value = 50,
                           min = 1,
                           step = 1))),
  fluidRow(
    column(4, numericInput(inputId = "b1",
                           label = HTML("b<sub>1</sub>: Half-saturation constant for sp. 1"),
                           value = 10,
                           min = 0,
                           step = 1)),
    column(4, numericInput(inputId = "b2",
                           label = HTML("b<sub>2</sub>: Half-saturation constant for sp. 2"),
                           value = 10,
                           min = 0,
                           step = 1)),
    column(2, radioButtons(inputId = "plot",
                           label = "Plot",
                           choices = c("N vs. time" = "time",
                                       "Phase portrait" = "phase",
                                       "Both" = "both"),
                           selected = "both")),
    column(2, conditionalPanel(
      condition = "input.plot != 'time'",
      checkboxGroupInput(inputId = "phase_opt",
                         label = "Draw",
                         choices = c("Vector field" = "vec",
                                     "Population trajectories" = "traj"),
                         selected = c("vec", "traj")))))
)

# Output panel
fluidRow(
  column(12, renderPlot(plot(), height = 500))
)

##### Server #####

# Validate input
observeEvent(input$a12, {
  if (input$a12 <= 0) updateNumericInput(inputId = "a12", value = 0.1)
})
observeEvent(input$a21, {
  if (input$a21 <= 0) updateNumericInput(inputId = "a21", value = 0.1)
})
observeEvent(input$d1, {
  if (input$d1 <= 0) updateNumericInput(inputId = "d1", value = 0.01)
})
observeEvent(input$d2, {
  if (input$d2 <= 0) updateNumericInput(inputId = "d2", value = 0.01)
})
observeEvent(input$b1, {
  if (input$b1 <= 0) updateNumericInput(inputId = "b1", value = 1)
})
observeEvent(input$b2, {
  if (input$b2 <= 0) updateNumericInput(inputId = "b2", value = 1)
})
observeEvent(input$N1, {
  if (input$N1 <= 0) updateNumericInput(inputId = "N1", value = 1)
})
observeEvent(input$N2, {
  if (input$N2 <= 0) updateNumericInput(inputId = "N2", value = 1)
})
observeEvent(input$time, {
  if (input$time < 1) updateNumericInput(inputId = "time", value = 1)
})

# Retrieve user inputs
params <- reactive({c(r1 = input$r1,
                      r2 = input$r2,
                      a12 = input$a12,
                      a21 = input$a21,
                      d1 = input$d1,
                      d2 = input$d2,
                      b1 = input$b1,
                      b2 = input$b2)})
init <- reactive({c(N1 = input$N1, N2 = input$N2)})
time <- reactive({input$time})
vec <- reactive({"vec" %in% input$phase_opt})
traj <- reactive({"traj" %in% input$phase_opt})

# Run mutualism model
sim <- reactive({run_mutualism(time(), init(), params())})

# Plot population trajectories over time
plot <- reactive({
  if (input$plot == "time") {
    plot_mutualism_time(sim()) +
      ggplot2::theme(aspect.ratio = 1)
  } else if (input$plot == "phase") {
    plot_mutualism_portrait(sim(), vec(), traj()) +
      ggplot2::theme(aspect.ratio = 1)
  } else {
    patchwork::wrap_plots(plot_mutualism_time(sim()),
                          plot_mutualism_portrait(sim(), vec(), traj()),
                          nrow = 1) +
      ggplot2::theme(aspect.ratio = 1)
  }
})
```

## References

- Bronstein, J.L. (2015). [*Mutualism*](https://doi.org/10.1093/acprof:oso/9780199675654.001.0001). Oxford University Press, Oxford, UK.
- Cook, J.M. & Rasplus, J.-Y. (2003). [Mutualists with attitude: coevolving fig wasps and figs](https://doi.org/10.1016/S0169-5347(03)00062-4). *Trends in Ecology and Evolution*, 18, 241-248.
- Janzen, D.H. (1966). [Coevolution between ants and acacias in Central America](https://doi.org/10.2307/2406628). *Evolution*, 20, 249-275.

```{r footer, results ='asis'}
# Print footer
suppressWarnings(print_app_footer())
```
